---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# marginplyr

### SQL-style grouping sets, rollups, and cubes for dplyr and dbplyr

Create detail rows, subtotals, and grand totals with one dplyr-style summary.
Use the same grouping specification with a local data frame or a lazy database
table.

<!-- badges: start -->
[![R-CMD-check](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml)
[![altdoc](https://github.com/sayuks/marginplyr/actions/workflows/altdoc.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/altdoc.yaml)
[![Codecov test coverage](https://codecov.io/gh/sayuks/marginplyr/graph/badge.svg)](https://app.codecov.io/gh/sayuks/marginplyr)
[![lint.yaml](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml)
<a href = "https://sayuks.github.io/marginplyr/" target = "_blank"><img src="https://cranlogs.r-pkg.org/badges/marginplyr"></a>
<!-- badges: end -->

```{r}
library(marginplyr)

january_sales <- retail_sales |>
  dplyr::filter(year == 2026L, month == "Jan")

january_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )
```

`rollup(region, store)` asks for three reporting levels:

```text
(region, store)  store detail
(region)         region subtotal
()               company total
```

The result is one regular data frame—not three reports that you have to build
and combine yourself.

- Works with local data frames and lazy `dbplyr` tables.

- Uses native `GROUP BY GROUPING SETS` on DuckDB and PostgreSQL.

- Uses a portable `UNION ALL` fallback for other SQL dialects.

- Distinguishes a source `NA` from a dimension removed to create a total.

- Exposes Grouping set identifiers and SQL-compatible absence masks without
  relying on display labels.

- Calculates each scalar summary's share of its immediate rollup parent.

- Can expand or nest the source rows behind every subtotal.

New to the idea? Start with [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html).
Need to compare `.id`, Grouping bits, and Grouping identifiers? See [Grouping
identity](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html).
Need absent dimension keys? See [Complete absent keys before
margins](https://sayuks.github.io/marginplyr/vignettes/completing_keys.html).
Already working with remote data? See [Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html).

## Installation

You can install `marginplyr` from CRAN:

``` r
install.packages("marginplyr")
```

You can install the development version from
[GitHub](https://github.com/sayuks/marginplyr) with:

``` r
install.packages("pak")
pak::pkg_install("sayuks/marginplyr")
```

## Grouping sets in plain language

A normal grouped summary has one grain: for example, one row per store. A
management report often needs several grains together: stores, regions, and
the whole company. SQL calls that collection of grains *grouping sets*.

marginplyr provides three ways to describe them:

| Helper | Use it when | For `region` and `store` |
|---|---|---|
| `rollup()` | The dimensions form a hierarchy | store detail, region subtotal, company total |
| `grouping_sets()` | You need an exact list of unrelated views | only the combinations you request |
| `cube()` | You need every combination | detail, both one-dimensional subtotals, grand total |

Columns in `.by` stay fixed in every result level. For example, this creates a
separate store-to-company report for each year and month:

```{r}
monthly_report <- retail_sales |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .by = c(year, month),
    .grouping = rollup(region, store)
  )

monthly_report |>
  dplyr::filter(year == 2026L, month == "Jan")
```

Use `.by` for report partitions that must never become totals, and
`.grouping` for dimensions that may be aggregated away.

## Ask for the exact views you need

`grouping_sets()` combines an explicit list of grouping levels. An empty
`grouping_set()` requests the grand total:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      grouping_set(year, month),
      grouping_set(region, product),
      grouping_set()
    )
  )
```

`cube()` creates every subset of its dimensions:

```{r}
january_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(product, channel)
  )
```

## A total is not the same thing as `NA`

Some rows in `retail_sales` have `store = NA` because an online-direct sale
does not belong to a physical store. That source value must remain different
from a region subtotal where `store` was deliberately removed.

```{r}
january_sales |>
  dplyr::filter(region == "East") |>
  summarize_with_margins(
    revenue = sum(revenue),
    store_is_total = grouping_bit(store),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )
```

`grouping_bit(store)` is `0` for the source `NA` row and `1` when the rollup
removed `store`. This is the role of the SQL `GROUPING()` function; replacing
every missing value with `"Total"` cannot preserve that distinction.

## Compare each summary with its rollup parent

`share_of_parent()` divides a preceding named summary by the same summary one
rollup level up. Name the source summary first, then take its share:

```{r}
parent_report <- january_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    revenue_share = share_of_parent(revenue),
    .grouping = rollup(region, store)
  )

parent_report
```

For several measures, use two ordered `across()` expressions. The second one
selects the summaries the first one created:

```{r}
january_sales |>
  summarize_with_margins(
    dplyr::across(c(units, revenue), sum),
    dplyr::across(
      c(units, revenue),
      share_of_parent,
      .names = "{.col}_share"
    ),
    .grouping = rollup(region, store)
  )
```

Percentages, rounding, and other derived columns belong in a `mutate()` after
the summary:

```{r}
parent_report |>
  dplyr::mutate(revenue_percent = 100 * revenue_share)
```

The root row's share is one, and a missing or zero denominator gives `NA`.
Parent matching is structural, so display labels never choose the parent. The
[`share_of_parent()`
reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
documents the eligible sources, value rules, `across()` grammar, and backend
boundaries in full.

## Use the same report with a database

The grouping plan becomes part of the lazy query. `show_query()` inspects the
SQL without collecting the result:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = rollup(region, store)
  ) |>
  dplyr::show_query()
```

DuckDB and PostgreSQL use native grouping sets. Backends without confirmed
native support use a `UNION ALL` translation with the same grouping plan.
Simulation verifies generated SQL but does not claim live execution against
every database server. The [database
guide](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
shows native SQL, fallback SQL, live DuckDB execution, `collect()`, and the
verification status of each backend.

## Keep the rows behind each total

SQL-style aggregation is only the start. `nest_with_margins()` keeps the
source records for every detail group, subtotal, and total in a list column:

```{r}
nested_sections <- january_sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

# Convert to a tibble so nested data frames print compactly.
nested_sections |>
  dplyr::as_tibble() |>
  head()

nested_sections$data[[1]]
```

Use `.keep = TRUE` to retain the original grouping columns inside each nested
table. Use `nest_by_with_margins()` when the next calculation should run
row-wise against each nested report section, or `expand_with_margins()` when
downstream work needs expanded rows rather than an immediate summary.

These are margin-aware counterparts, not drop-in replacements for
`tidyr::nest()` and `dplyr::nest_by()`. Their precise grouping, `.keep`,
empty-input, and lazy-backend contracts are documented in the function
reference and Get Started guide.

## Choose the tool for the job

These approaches solve related problems with different interfaces:

| Approach | Best fit | Result and execution model |
|---|---|---|
| Repeated `dplyr::summarize()` plus row binding | A small, one-off set of totals where explicit branches are clearest | You author and combine each grouping level |
| [`data.table`](https://rdatatable.gitlab.io/data.table/reference/groupingsets.html) grouping sets | A local data.table workflow | `rollup()`, `cube()`, and `groupingsets()` return one data.table |
| [`rollup`](https://cran.r-project.org/package=rollup) | Its dplyr-oriented grouped-data-frame-list workflow fits the analysis | Grouping levels are represented as a list before their summaries are combined |
| marginplyr | The same grouping plan should work locally and on a lazy source | One data frame or lazy query, with native SQL or a portable fallback |

marginplyr is most useful when a report has multiple grains, when the data
should stay in a database until `collect()`, or when subtotal identity and the
records behind each total matter. For one simple local subtotal, explicit
dplyr code may be easier.

## Backend verification

| Verification status | Backends |
|---|---|
| Live native execution tested | DuckDB |
| Live portable Parent-share execution tested | SQLite |
| Native SQL generation tested | PostgreSQL |
| Fallback SQL generation tested | Access, SAP HANA, Hive, Impala, MariaDB, Microsoft SQL Server, MySQL, Oracle, Amazon Redshift, Snowflake, Spark SQL, SQLite, Teradata, and generic DBI/ODBC connections |
| Non-SQL lazy backend tested | Arrow, dtplyr |

The function reference contains executable examples for composite dimensions,
tidy-select expressions, duplicate grouping sets, Cartesian products,
computed grouping columns, simulated SQL dialects, DuckDB, and dtplyr.

## Code of Conduct

Please note that the marginplyr project is released with a [Contributor Code
of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
