---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# marginplyr

<!-- badges: start -->
[![R-CMD-check](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml)
[![altdoc](https://github.com/sayuks/marginplyr/actions/workflows/altdoc.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/altdoc.yaml)
[![Codecov test coverage](https://codecov.io/gh/sayuks/marginplyr/graph/badge.svg)](https://app.codecov.io/gh/sayuks/marginplyr)
[![lint.yaml](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml)
<a href = "https://sayuks.github.io/marginplyr/" target = "_blank"><img src="https://cranlogs.r-pkg.org/badges/marginplyr"></a>
<!-- badges: end -->

`marginplyr` brings SQL-style `GROUPING SETS`, `ROLLUP`, and `CUBE` to
`dplyr`. The same grouping specification works with local data frames and
lazy database tables, with `"Total"` as the default display label.

## Installation

<!-- You can install `marginplyr` from CRAN: -->

<!-- ``` r -->
<!-- install.packages("marginplyr") -->
<!-- ``` -->
You can install the development version of `marginplyr` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("sayuks/marginplyr")
```

## From a monthly report to a database query

```{r}
suppressPackageStartupMessages(library(marginplyr))
```

Imagine an analyst preparing a management report for a retail chain. Each row
of `retail_sales` records sales for one product, channel, and store
combination. Online-direct sales have no physical store, so `store` is
genuinely missing in some source rows.

```{r}
dplyr::slice_head(retail_sales, n = 6)
```

### Store, region, and company totals

The first request is one independent report per month. `year` and `month`
belong to every result level, while `rollup(region, store)` creates store
detail, region subtotals, and a company total.

```{r}
monthly_report <- summarize_with_margins(
  retail_sales,
  units = sum(units),
  revenue = sum(revenue),
  .by = c(year, month),
  .grouping = rollup(region, store)
)

monthly_report |>
  dplyr::filter(year == 2026L, month == "Jan")
```

`"Total"` means that a dimension was aggregated away. A source `NA` is left
as `NA`; it still represents an online-direct record with no physical store.

Moving the period columns into the rollup changes the report hierarchy:

```{r}
all_levels <- summarize_with_margins(
  retail_sales,
  revenue = sum(revenue),
  level = grouping_id(year, month, region, store),
  .grouping = rollup(year, month, region, store)
)

all_levels |>
  dplyr::filter(month == "Total")
```

This rollup includes store detail, region-month subtotals, company-month
totals, company-year totals, and the all-period total.

### Ask for exactly the totals you need

Finance does not always want a hierarchy. `grouping_sets()` selects arbitrary
views, and an empty `grouping_set()` requests the grand total.

```{r}
summarize_with_margins(
  retail_sales,
  revenue = sum(revenue),
  .grouping = grouping_sets(
    grouping_set(year, month),
    grouping_set(region, product),
    grouping_set()
  )
)
```

Merchandising can instead explore every combination of product and sales
channel with `cube()`:

```{r}
retail_sales |>
  dplyr::filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(product, channel)
  )
```

### Tell source missing values from totals

SQL uses `GROUPING()` because both a source missing value and an aggregated
dimension would otherwise appear as `NULL`. The marginplyr helpers expose the
same distinction:

```{r}
retail_sales |>
  dplyr::filter(
    year == 2026L,
    month == "Jan",
    region == "East"
  ) |>
  summarize_with_margins(
    revenue = sum(revenue),
    store_is_total = grouping_bit(store),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )
```

`store_is_total` is `0` for the source `NA` row and `1` for a subtotal where
the store dimension was removed.

### Backend support

marginplyr separates native grouping-set support from its portable
`UNION ALL` fallback:

| Execution path | Backends | Automated verification |
|---|---|---|
| Native `GROUP BY GROUPING SETS` | DuckDB, PostgreSQL | Live DuckDB execution; PostgreSQL SQL simulation |
| Portable `UNION ALL` | Access, SAP HANA, Hive, Impala, MariaDB, Microsoft SQL Server, MySQL, Oracle, Amazon Redshift, Snowflake, Spark SQL, SQLite, Teradata, and generic DBI/ODBC connections | SQL generation with dbplyr simulators |

Arrow and dtplyr are also tested lazy backends, but they are not SQL database
connections. Simulated coverage verifies generated SQL; it does not claim
live execution against every listed database server.

### Run the same report in DuckDB

Local data frames and lazy database tables use the same interface. DuckDB and
PostgreSQL translate the request to one native `GROUP BY GROUPING SETS` query.
Install the optional backend packages before running this example:

``` r
install.packages(c("DBI", "duckdb"))
```

```{r duckdb}
con <- suppressMessages(DBI::dbConnect(duckdb::duckdb()))

sales_db <- dplyr::copy_to(
  con,
  retail_sales,
  name = "retail_sales",
  temporary = TRUE,
  overwrite = TRUE
)

query <- sales_db |>
  dplyr::filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store),
    .sort = TRUE
  )

dplyr::show_query(query)
dplyr::collect(query)

DBI::dbDisconnect(con)
```

Backends without confirmed native support use a portable `UNION ALL`
translation with the same grouping semantics.

### Keep the rows behind each total

`union_all_with_margins()` expands source rows across grouping levels.
`nest_by_with_margins()` goes one step further and keeps a row-wise detail
table behind every visible group:

```{r}
retail_sales |>
  dplyr::filter(year == 2026L, month == "Jan") |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  ) |>
  dplyr::mutate(
    records = nrow(data),
    revenue = sum(data$revenue)
  ) |>
  dplyr::select(-data)
```

The function reference contains executable examples for composite dimensions,
tidy-select expressions, duplicate grouping sets, Cartesian products,
simulated SQL dialects, and dtplyr.

The [Get started
guide](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
develops the complete story, including the difference between the union made
by `grouping_sets()` and the Cartesian product made by `grouping_spec()`, plus
a small [`quartabs`](https://sayuks.github.io/quartabs/) reporting example.

## Code of Conduct

Please note that the marginplyr project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
