---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# marginplyr

<!-- badges: start -->
<a href = "https://CRAN.R-project.org/package=marginplyr" target = "_blank"><img src="https://www.r-pkg.org/badges/version/marginplyr"></a>
[![R-CMD-check](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sayuks/marginplyr/graph/badge.svg)](https://app.codecov.io/gh/sayuks/marginplyr)
[![lint.yaml](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml/badge.svg)](https://github.com/sayuks/marginplyr/actions/workflows/lint.yaml)
<a href = "https://sayuks.github.io/marginplyr/" target = "_blank"><img src="https://cranlogs.r-pkg.org/badges/marginplyr"></a>
<!-- badges: end -->

`marginplyr` brings SQL-style `GROUPING SETS`, `ROLLUP`, and `CUBE` to
`dplyr`. The same grouping specification works with local data frames and
lazy database tables, with `"Total"` as the default display label.

## Installation

<!-- You can install `marginplyr` from CRAN: -->

<!-- ``` r -->
<!-- install.packages("marginplyr") -->
<!-- ``` -->
You can install the development version of `marginplyr` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("sayuks/marginplyr")
```

## Example

```{r}
library(marginplyr)
```

### Hierarchical totals with `rollup()`

```{r}
summarize_with_margins(
  mtcars,
  n = dplyr::n(),
  mean_mpg = mean(mpg, na.rm = TRUE),
  level = grouping_id(cyl, vs),
  .grouping = rollup(cyl, vs)
)
```

### Arbitrary grouping sets, including the grand total

```{r}
summarize_with_margins(
  mtcars,
  n = dplyr::n(),
  .grouping = grouping_sets(
    grouping_set(cyl, vs),
    grouping_set(cyl, gear),
    grouping_set()
  )
)
```

`grouping_spec()` combines grouping families by Cartesian product. Composite
dimensions stay together inside a rollup or cube:

```{r}
summarize_with_margins(
  mtcars,
  n = dplyr::n(),
  .grouping = grouping_spec(
    rollup(cyl, vs),
    cube(grouping_set(gear, carb), am)
  )
) |>
  dplyr::slice_head(n = 12)
```

DuckDB and PostgreSQL use one native `GROUP BY GROUPING SETS` query. Other
lazy backends use a portable `UNION ALL` adapter with the same result
semantics. `grouping()` and `grouping_id()` distinguish source missing values
from rows produced by aggregation.

See [Get started](https://sayuks.github.io/marginplyr/vignettes/get_started.html) for more details.

## Code of Conduct

Please note that the marginplyr project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
