---
title: Complete absent keys before margins
vignette: >
  %\VignetteIndexEntry{Complete absent keys before margins}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

marginplyr creates the Grouping sets requested by a Grouping specification; it
does not create unobserved values within those sets. Complete absent keys
before a Margin operation when synthetic facts should participate in detail,
subtotal, and grand-total summaries. Replace values after a summary when the
change is only for display.

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. The tidyr and SQLite sections below need
optional packages, so their chunks evaluate only when those packages are
installed at the versions marginplyr asks for. Where a package is missing or
too old the code is still shown, but its output is absent.

```{r}
#| include: false

# `marginplyr_suggest_available()` answers what a bare installation check
# could not: whether the installed package satisfies the version DESCRIPTION
# requires. AGENTS.md is authoritative for which guards must go through it.
source(system.file("suggests", "guard.R", package = "marginplyr"))

has_tidyr <- marginplyr_suggest_available("tidyr")
has_sqlite <- marginplyr_suggest_available("DBI") &&
  marginplyr_suggest_available("RSQLite")
```

## Why completion is a separate operation

marginplyr does not provide `.complete`, `.fill`, or `.empty` arguments, and it
does not provide `complete_with_margins()`.

Input completion and result completion answer different questions:

- **Pre-summary synthetic facts** participate in every requested Grouping set.
  They can change counts, means, distinct counts, and user-defined summaries.
- **Post-summary display replacement** changes how an existing result is
  shown. It does not recalculate subtotals or the grand total.

Sums can often be reconstructed from child sums, but arbitrary summaries
cannot. A mean needs weights, distinct counts can overlap, and medians cannot
generally be combined from child medians. Keeping completion outside the
Margin operation makes the chosen semantics visible in ordinary data
manipulation code.

Completion also does not belong in a Grouping specification. A Grouping
specification declares which Grouping sets exist; it does not declare the
valid value domain of each dimension or invent fact values.

## Complete local facts with `tidyr::complete()`

This small fact table does not contain every product for every observed
region-store relationship:

```{r}
facts <- tibble(
  fact_id = 1:3,
  region = c("East", "East", "West"),
  store = c("A", "B", "C"),
  product = c("Laptop", "Monitor", "Laptop"),
  units = c(2L, 3L, 4L),
  revenue = c(200, 300, 400)
)

product_domain <- c("Laptop", "Monitor")
facts
```

`tidyr::nesting(region, store)` preserves only observed region-store
relationships while crossing each relationship with the product domain:

```{r}
#| eval: !expr has_tidyr

completed_facts <- facts |>
  mutate(is_fact = TRUE) |>
  tidyr::complete(
    tidyr::nesting(region, store),
    product = product_domain,
    fill = list(
      is_fact = FALSE,
      units = 0L,
      revenue = 0
    )
  )

completed_facts
```

The inserted zeros now participate in every level of the rollup:

```{r}
#| eval: !expr has_tidyr

completed_facts |>
  summarize_with_margins(
    rows = dplyr::n(),
    observed_facts = sum(is_fact),
    units = sum(units),
    mean_per_completed_key = mean(revenue),
    mean_per_observed_fact = mean(revenue[is_fact]),
    .grouping = rollup(region, store, product),
    .id = "occurrence"
  ) |>
  arrange(occurrence, region, store, product)
```

`rows` counts completed keys, while `observed_facts` counts only source facts.
The retained `fact_id` and `is_fact` marker make that distinction explicit.
The two means show that filling synthetic facts before a summary is a
substantive modeling choice.

If the intended change is only presentational, summarize the observed facts
first and replace the finished display value afterwards:

```{r}
observed_summary <- facts |>
  summarize_with_margins(
    mean_revenue = mean(revenue),
    .grouping = rollup(region, store, product)
  )

observed_summary |>
  mutate(display_mean = dplyr::coalesce(mean_revenue, 0))
```

That `mutate()` does not add fact rows or alter any existing aggregate.

## Union an explicit scaffold with observed keys

Sometimes the requested domain is an explicit relation rather than a regular
cross product. Union it with observed keys before joining so facts outside the
scaffold are not accidentally discarded:

```{r}
requested_keys <- tribble(
  ~region, ~store, ~product,
  "East", "A", "Laptop",
  "East", "A", "Monitor",
  "West", "C", "Laptop",
  "West", "C", "Monitor"
)

all_keys <- requested_keys |>
  union(distinct(facts, region, store, product))

scaffolded_facts <- all_keys |>
  left_join(facts, by = c("region", "store", "product")) |>
  mutate(
    is_fact = !is.na(fact_id),
    units = dplyr::coalesce(units, 0L),
    revenue = dplyr::coalesce(revenue, 0)
  )

scaffolded_facts
```

The `left_join()` makes one synthetic fact for every scaffold key without an
observed fact. Keep a fact identifier or marker whenever counts, means, or
other summaries must distinguish those rows.

## Complete keys on a database connection

Use ordinary lazy relations on the same connection. An existing dimension
table is the natural source of a reusable key domain:

```{r}
#| message: false
#| eval: !expr has_sqlite

con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")

facts_db <- copy_to(
  con,
  facts,
  name = "facts",
  temporary = TRUE,
  overwrite = TRUE
)
dimension_db <- copy_to(
  con,
  requested_keys,
  name = "store_product_dimension",
  temporary = TRUE,
  overwrite = TRUE
)

observed_keys_db <- facts_db |>
  distinct(region, store, product)
all_keys_db <- dimension_db |>
  select(region, store, product) |>
  union(observed_keys_db)

completed_db <- all_keys_db |>
  left_join(
    facts_db,
    by = c("region", "store", "product")
  ) |>
  mutate(
    is_fact = !is.na(fact_id),
    units = dplyr::coalesce(units, 0L),
    revenue = dplyr::coalesce(revenue, 0)
  )

completed_query <- completed_db |>
  summarize_with_margins(
    observed_facts = sum(is_fact),
    revenue = sum(revenue),
    .grouping = rollup(region, store, product)
  )

show_query(completed_query)
DBI::dbDisconnect(con)
```

The union, join, completion expressions, and Margin summary remain lazy.
`show_query()` renders the staged SQL without executing the finished report.

## Move local keys only when you ask to

For a small local key table, `dbplyr::copy_inline()` embeds the rows in the
lazy SQL with `VALUES`. This avoids creating a persistent table but makes the
query text grow with the number of keys:

```{r}
#| message: false
#| eval: !expr has_sqlite

con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
facts_db <- copy_to(
  con,
  facts,
  name = "facts",
  temporary = TRUE,
  overwrite = TRUE
)

inline_keys <- dbplyr::copy_inline(con, requested_keys)
inline_query <- inline_keys |>
  union(distinct(facts_db, region, store, product)) |>
  left_join(
    facts_db,
    by = c("region", "store", "product")
  )
show_query(inline_query)

reusable_keys <- copy_to(
  con,
  requested_keys,
  name = "requested_keys",
  temporary = TRUE,
  overwrite = TRUE
)
reusable_query <- reusable_keys |>
  union(distinct(facts_db, region, store, product)) |>
  left_join(
    facts_db,
    by = c("region", "store", "product")
  )
show_query(reusable_query)

DBI::dbDisconnect(con)
```

Use explicit `dplyr::copy_to()` for a larger or reusable local key relation.
It performs data movement at the line where the caller requested it.

marginplyr never copies, collects, uploads, downloads, or otherwise moves
completion data implicitly. Connection choice, data movement, and execution
remain ordinary dplyr and dbplyr operations owned by the caller.
