---
title: Database and lazy backends
vignette: >
  %\VignetteIndexEntry{Database and lazy backends}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` applies one grouping plan to local data frames and supported lazy
tables. With a database table, the plan remains lazy: `show_query()` renders
the SQL, while `collect()` executes it and brings the result into R.

If grouping sets are new to you, read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
first. This guide focuses on query shape, execution, and backend support.

## Installation

Install the development version of marginplyr from GitHub:

```{r}
#| eval: false

install.packages("pak")
pak::pkg_install("sayuks/marginplyr")

# Used for the optional backend sections of this guide.
install.packages(c("DBI", "duckdb", "RSQLite", "dtplyr"))
```

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. The DuckDB, SQLite, and dtplyr sections
below need optional packages, so their chunks evaluate only when those
packages are installed. Where a package is missing the code is still shown,
but its output is absent.

```{r}
#| include: false

has_duckdb <- requireNamespace("DBI", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE)
# dbplyr's SQLite simulator opens no connection, but rendering its SQL reads
# the installed driver version through RSQLite.
has_sqlite_simulator <- requireNamespace("RSQLite", quietly = TRUE)
has_dtplyr <- requireNamespace("dtplyr", quietly = TRUE)
```

## One report, two execution locations

The local report uses store detail, region subtotals, and a company total:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

local_result <- january_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

local_result
```

The grouping specification does not change when the input becomes a lazy table:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_postgres()
)

postgres_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

postgres_query
```

Printing a lazy table may show a preview, depending on the backend.
`show_query()` is the deliberate way to inspect the query without collecting
the result:

```{r}
show_query(postgres_query)
```

Use `show_query()` during interactive analysis. Use
`dbplyr::sql_render()` when code needs the rendered SQL value for logging,
testing, or another programmatic purpose.

## Native grouping sets

DuckDB and PostgreSQL use marginplyr's native path. The generated query
contains one `GROUP BY GROUPING SETS` operation. The requested rollup:

```text
(region, store)
(region)
()
```

becomes the equivalent SQL grouping-set list. Hidden `GROUPING()` flags allow
the outer query to label only dimensions removed by aggregation. A source
`NULL` therefore stays different from a generated subtotal even though SQL
would otherwise display both as missing.

More complex grouping specifications use the same translation. The two
rollups below are combined by Cartesian product, matching comma-separated SQL
`GROUP BY` items:

```{r}
dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
) |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  ) |>
  show_query()
```

`grouping_sets(rollup(...), rollup(...))` has different semantics: it forms
the union of two independent hierarchies. See [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html#union-versus-cartesian-product)
for the complete comparison.

## Execute in DuckDB

The simulator above verifies SQL rendering without a server. DuckDB lets the
same workflow execute against a live in-memory database:

```{r}
#| eval: !expr has_duckdb

# `shared_home = FALSE` keeps DuckDB's extension cache and stored secrets in
# the session's temporary directory rather than in `~/.duckdb`.
con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))

sales_db <- copy_to(
  con,
  retail_sales,
  name = "retail_sales",
  temporary = TRUE,
  overwrite = TRUE
)

duckdb_query <- sales_db |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

show_query(duckdb_query)
```

At this point `duckdb_query` is still lazy. `collect()` asks DuckDB to execute
the grouping sets and transfers only the summarized result:

```{r}
#| eval: !expr has_duckdb

duckdb_result <- collect(duckdb_query)
DBI::dbDisconnect(con)

duckdb_result
```

Use `compute()` instead when the next stage should remain in the database as a
materialized or temporary remote table.

DuckDB also preserves factor grouping columns represented as enums. When a
factor dimension uses `.margin_label = NA_character_` or `NULL`, marginplyr
leaves the omitted dimension as a typed SQL `NULL`; it never attempts to add
an NA enum label. A non-missing label is added as an enum level and is placed
last by default or first with `.margin_label_position = "first"`.

## Portable fallback SQL

marginplyr does not assume that every SQL dialect supports native grouping
sets. For an unconfirmed backend, it expands the plan into one query branch
per grouping set and joins those branches with `UNION ALL`.

```{r}
#| eval: !expr has_sqlite_simulator

sqlite_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
)

sqlite_query <- sqlite_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )

show_query(sqlite_query)
```

This SQL is longer than the native form, but it preserves the same requested
grouping levels and margin labels. Native versus fallback translation is
selected by the backend adapter; analysis code does not need a different
grouping specification.

`expand_with_margins()` also stays lazy. On a fallback backend, its SQL makes
the row-expansion branches explicit:

```{r}
#| eval: !expr has_sqlite_simulator

sqlite_sales |>
  expand_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

## Parent shares use a staged lazy query

`share_of_parent()` needs a denominator from another Grouping-set row, so it
cannot be translated as one ordinary aggregate expression. marginplyr first
forms the ordinary Margin summaries, then builds one Parent mapping shared by
every requested measure:

```{r}
parent_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_share = share_of_parent(revenue),
    .grouping = rollup(region, store)
  )

show_query(parent_query)
```

This is still one lazy query. `show_query()` renders it without execution;
only an explicit operation such as `collect()` or `compute()` asks the backend
to run it. Native and portable grouping adapters both feed the same staged
Parent-share contract. Parent matching uses internal Grouping identity and
missing-safe keys rather than displayed Margin labels.

`share_of_total()` stages the same way and asks less of the query. Its
denominator depends on the fixed `.by` keys and nothing else, so its mapping
is one filtered read of a row of the Grand total set, joined on those keys
alone — and on a constant when, as here, there are none:

```{r}
postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_of_total = share_of_total(revenue),
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

Syntax, source-name, written-order, and `across()` errors are validated before
execution. A share source must still be a plain integer or double, and that
rule holds here too. A general dbplyr backend exposes no metadata for an
arbitrary aggregate's result type, so marginplyr reads the summaries a share
reads over a single input row and rejects an ineligible source before
returning the query. That read is bounded in the rows it asks for and gets
back, not in the work the input may have to do to produce one row — the staged
query above is still not executed, and `show_query()` runs nothing. A source
the sample cannot type is left alone rather than rejected, so a runtime-only
incompatibility, and any non-scalar summary, can still surface from the
database when the staged query executes.

The portable value contract covers finite numbers, missing values, and zero
denominators. Backend-specific non-finite representations are outside that
contract. See the
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
for both helpers' supported sources, denominator rules, `across()` forms,
empty inputs, and post-summary `mutate()` guidance.

## Completion stays in the lazy input pipeline

marginplyr does not create absent dimension keys or move completion data
between R and a database. Use a same-connection dimension relation, or move a
local relation explicitly before the Margin operation. The [Complete absent
keys before margins
guide](https://sayuks.github.io/marginplyr/vignettes/completing_keys.html)
shows a lazy dimension-table union and join, `dbplyr::copy_inline()` for small
local keys, and explicit `dplyr::copy_to()` for larger or reusable keys.

## Other lazy backends

dtplyr is lazy without being a SQL connection. `show_query()` displays the
generated data.table expression:

```{r}
#| eval: !expr has_dtplyr

dt_query <- january_sales |>
  dtplyr::lazy_dt() |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

show_query(dt_query)
collect(dt_query)
```

`nest_with_margins()` also supports dtplyr and remains lazy until collection.
SQL database and Arrow tables cannot use the nesting verbs because their
result contains a list column.

dtplyr also enforces the contextual-share source rules, for both denominators.
A valid request stays a native lazy query, and an ineligible source type or a
non-scalar summary raises a Package condition at explicit execution, before an
invalid grouping row is emitted.

Arrow summaries and expansions are supported and lazy, but `share_of_parent()`
and `share_of_total()` are both rejected outright, before any Arrow query is
constructed: the reason is the source summary they share. Collect the Arrow
table first when you need either. The
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
states where each backend reports these errors.

## Backend verification levels

Support is not one status. SQL simulation is useful evidence about rendering,
but it is not the same as running a query on a live server, so it is worth
knowing which claim rests on which.

Only DuckDB and SQLite are exercised as live SQL databases. DuckDB covers
native SQL end to end — generation, execution, result values, labels, and
grouping identity — while SQLite's live claim is narrower: portable
`UNION ALL` execution for Parent shares, including missing-safe matching and
Grouping set identifiers. PostgreSQL and the dozen-plus fallback dialects are
verified through dbplyr simulators, which prove what SQL is rendered and
nothing about a running server. Arrow and dtplyr are tested for the lazy
behavior each supports.

Treat a simulator-only backend as SQL-generation support, and report
server-specific behavior with a minimal query and the backend version.
*Database backend coverage* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
names every verified backend.

## Practical rules for remote analysis

- Build the complete lazy pipeline before calling `collect()`.

- Use `show_query()` to confirm native `GROUPING SETS` or fallback
  `UNION ALL`.

- Use `.by` for columns that remain fixed in every grouping level.

- Keep `grouping_bit()` or `grouping_id()` when source missing values or the
  literal label `"Total"` can occur in the data.

- Lazy results have no guaranteed order unless the query explicitly requests
  one. Ask for a Margin order with `.sort`, or call `arrange()` for any other
  presentation order.

- Lazy inputs default to `.check_margin_label = FALSE` because detecting an
  existing label would require an additional query. Opt in only when that
  extra scan is appropriate.

- A fallback query may repeat the source relation once per grouping set.
  Inspect the generated SQL and the database query plan for large workloads;
  marginplyr does not claim that one translation is universally faster.
