---
title: Database and lazy backends
vignette: >
  %\VignetteIndexEntry{Database and lazy backends}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` applies one grouping plan to local data frames and supported lazy
tables. With a database table, the plan remains lazy: `show_query()` renders
the SQL, while `collect()` executes it and brings the result into R.

If grouping sets are new to you, read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
first. This guide focuses on query shape, execution, and backend support.

## Installation

Install marginplyr from GitHub:

```{r}
#| eval: false

install.packages("pak")
pak::pkg_install("sayuks/marginplyr")

# Used for the optional backend sections of this guide.
install.packages(c("DBI", "duckdb", "RSQLite", "dtplyr", "arrow"))
```

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. The DuckDB, SQLite, dtplyr, and Arrow
sections below need optional packages, so their chunks evaluate only when those
packages are installed at the versions marginplyr asks for. Where a package is
missing or too old the code is still shown, but its output is absent.

```{r}
#| include: false

# `marginplyr_suggest_available()` answers what a bare installation check
# could not: whether the installed package satisfies the version DESCRIPTION
# requires. AGENTS.md is authoritative for which guards must go through it.
source(system.file("suggests", "guard.R", package = "marginplyr"))

has_duckdb <- marginplyr_suggest_available("DBI") &&
  marginplyr_suggest_available("duckdb")
# dbplyr's SQLite simulator opens no connection, but rendering its SQL reads
# the installed driver version through RSQLite.
has_sqlite_simulator <- marginplyr_suggest_available("RSQLite")
has_dtplyr <- marginplyr_suggest_available("dtplyr")
has_arrow <- marginplyr_suggest_available("arrow")

# Chunks marked with `must_error` show a call that is meant to fail, so the
# reader sees the real diagnostic rather than a quotation of one. AGENTS.md is
# authoritative for what the option does and why it exists.
source(system.file("vignette-hooks", "must-error.R", package = "marginplyr"))
```

## One report, two execution locations

The local report uses store detail, region subtotals, and a company total:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

local_result <- january_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

local_result
```

The grouping specification does not change when the input becomes a lazy table:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_postgres()
)

postgres_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

postgres_query
```

Printing a lazy table may show a preview, depending on the backend.
`show_query()` is the deliberate way to inspect the query without collecting
the result:

```{r}
show_query(postgres_query)
```

Use `show_query()` during interactive analysis. Use
`dbplyr::sql_render()` when code needs the rendered SQL value for logging,
testing, or another programmatic purpose.

## Native grouping sets

DuckDB and PostgreSQL use marginplyr's native path. The generated query
contains one `GROUP BY GROUPING SETS` operation. The requested rollup:

```text
(region, store)
(region)
()
```

becomes the equivalent SQL grouping-set list. Hidden `GROUPING()` flags allow
the outer query to label only dimensions removed by aggregation. A source
`NULL` therefore stays different from a generated subtotal even though SQL
would otherwise display both as missing.

More complex grouping specifications use the same translation. The two
rollups below are combined by Cartesian product, matching comma-separated SQL
`GROUP BY` items:

```{r}
dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
) |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  ) |>
  show_query()
```

`grouping_sets(rollup(...), rollup(...))` has different semantics: it forms
the union of two independent hierarchies. See [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html#union-versus-cartesian-product)
for the complete comparison.

## Execute in DuckDB

The simulator above verifies SQL rendering without a server. DuckDB lets the
same workflow execute against a live in-memory database:

```{r}
#| eval: !expr has_duckdb

# `shared_home = FALSE` keeps DuckDB's extension cache and stored secrets in
# the session's temporary directory rather than in `~/.duckdb`.
con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))

sales_db <- copy_to(
  con,
  retail_sales,
  name = "retail_sales",
  temporary = TRUE,
  overwrite = TRUE
)

duckdb_query <- sales_db |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)

show_query(duckdb_query)
```

At this point `duckdb_query` is still lazy. `collect()` asks DuckDB to execute
the grouping sets and transfers only the summarized result:

```{r}
#| eval: !expr has_duckdb

duckdb_result <- collect(duckdb_query)
DBI::dbDisconnect(con)

duckdb_result
```

Use `compute()` instead when the next stage should remain in the database as a
materialized or temporary remote table.

DuckDB also preserves factor grouping columns represented as enums. When a
factor dimension uses `.margin_label = NA_character_` or `NULL`, marginplyr
leaves the omitted dimension as a typed SQL `NULL`; it never attempts to add
an NA enum label. A non-missing label is added as an enum level and is placed
last by default or first with `.margin_label_position = "first"`.

## `.check_margin_label` on lazy inputs

Lazy inputs default to `.check_margin_label = FALSE`, and what that default
gives up is not free: an *observed* collision — a real value already in the
data equal to `.margin_label` — goes undetected, so the result carries one
row for that value and one row for the margin, and neither the row itself nor
code reading the result afterward can tell which is which. *When marginplyr
queries your data* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
states the rule this default follows and its two exemptions.

Two protections hold regardless of `.check_margin_label`, at no added query.

A label equal to a *declared* factor level is rejected on every backend,
because the level is already known from the column's metadata and finding
the collision reads nothing. The enum preservation above is what makes this
visible on DuckDB: `region` below declares `"Total"` as an unused level, and
`.margin_label = "Total"` is rejected before any query is built, even though
`.check_margin_label` is left at its lazy default:

```{r}
#| eval: !expr has_duckdb

con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
region_data <- data.frame(
  region = factor(c("East", "West"), levels = c("East", "West", "Total")),
  revenue = c(10, 20)
)
region_db <- copy_to(
  con,
  region_data,
  name = "region_data",
  temporary = TRUE,
  overwrite = TRUE
)
```

```{r}
#| label: declared-level-rejected
#| eval: !expr has_duckdb
#| must_error: marginplyr_error

region_db |>
  summarize_with_margins(
    total = sum(revenue),
    .grouping = rollup(region)
  )
```

```{r}
#| eval: !expr has_duckdb

DBI::dbDisconnect(con)
```

Keeping `grouping_bit()` or `grouping_id()` in the result is the other free
protection: either identifies every margin row structurally, so a source row
and a margin row that print alike still stay distinguishable by position,
whatever `.check_margin_label` finds or misses.

Opt into `.check_margin_label = TRUE` when the extra scan for an observed
collision is worth the query for your backend and workload.

## Portable fallback SQL

marginplyr does not assume that every SQL dialect supports native grouping
sets. For an unconfirmed backend, it expands the plan into one query branch
per grouping set and joins those branches with `UNION ALL`.

```{r}
#| eval: !expr has_sqlite_simulator

sqlite_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
)

sqlite_query <- sqlite_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )

show_query(sqlite_query)
```

This SQL is longer than the native form, but it preserves the same requested
grouping levels and margin labels. Native versus fallback translation is
selected by the backend adapter; analysis code does not need a different
grouping specification.

`expand_with_margins()` also stays lazy. On a fallback backend, its SQL makes
the row-expansion branches explicit:

```{r}
#| eval: !expr has_sqlite_simulator

sqlite_sales |>
  expand_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

## Parent shares use a staged lazy query

`share_of_parent()` needs a denominator from another Grouping-set row, so it
cannot be translated as one ordinary aggregate expression. marginplyr first
forms the ordinary Margin summaries, then builds one Parent mapping shared by
every requested measure. `.check_share_source = FALSE` appears here because
the simulated connection executes nothing and therefore cannot be asked what
its dialect does with an ineligible source; the rule it opts out of is
described below.

```{r}
parent_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_share = share_of_parent(revenue),
    .grouping = rollup(region, store),
    .check_share_source = FALSE
  )

show_query(parent_query)
```

This is still one lazy query. `show_query()` renders it without execution;
only an explicit operation such as `collect()` or `compute()` asks the backend
to run it. Native and portable grouping adapters both feed the same staged
Parent-share contract. Parent matching uses internal Grouping identity and
missing-safe keys rather than displayed Margin labels.

`share_of_total()` stages the same way and asks less of the query. Its
denominator depends on the fixed `.by` keys and nothing else, so its mapping
is one filtered read of a row of the Grand total set, joined on those keys
alone — and on a constant when, as here, there are none:

```{r}
postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_of_total = share_of_total(revenue),
    .grouping = rollup(region, store),
    .check_share_source = FALSE
  ) |>
  show_query()
```

Syntax, source-name, written-order, and `across()` errors are validated before
execution. A share source must still be a plain integer or double, and that
rule holds here too — but marginplyr reads no row of your data to establish
it. A database evaluates the source summary itself, so where the dialect
refuses an ineligible summary that refusal is the answer, and it reaches you
as the database's own diagnostic when you execute the query. The staged
queries above are still not executed, and `show_query()` runs nothing.

Where the dialect converts a value of another type to a number instead of
refusing it — SQLite answers `sum()` of a text column with `0` rather than an
error — nothing applies the rule at all, and marginplyr refuses the share
rather than calculating one from values nothing has checked. Which of the two
a dialect does is asked with at most two queries referencing none of your
tables: a probe, and — only when the probe is rejected — a control, which is
what tells a dialect that genuinely refuses apart from one that could not
answer at all. An answer is a property of the dialect, so a dialect that
answers is asked once and every later connection carrying it reuses that
answer. Where the question could not be answered, nothing is remembered and
the next share request asks again, so a dropped connection or a warehouse that
was resuming does not refuse your shares for the rest of the session. A
connection that executes nothing, such as the simulators used above, cannot be
asked and is refused for the same reason; `.check_share_source = FALSE`
calculates the share from sources you have established yourself.

A refusing dialect is therefore shown against a live DuckDB table rather than
the simulator used above. The diagnostic is DuckDB's own, and the column it
names carries the name of the summary to rewrite. It reaches you as a plain
`rlang_error` with no marginplyr class in its `parent` chain — nothing here
reads the row that used to make it a `marginplyr_error` — so the chunk below
accepts any error rather than naming one:

```{r}
#| eval: !expr has_duckdb

sample_con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
sample_sales <- copy_to(
  sample_con,
  january_sales,
  name = "january_sales",
  temporary = TRUE,
  overwrite = TRUE
)
```

```{r}
#| label: duckdb-ineligible-share-source
#| eval: !expr has_duckdb
#| must_error: true

sample_sales |>
  summarize_with_margins(
    last_channel = max(channel),
    last_channel_share = share_of_parent(last_channel),
    .grouping = rollup(region, store)
  ) |>
  collect()
```

```{r}
#| eval: !expr has_duckdb

DBI::dbDisconnect(sample_con)
```

Cardinality is not established this way at all: a SQL aggregate returns one
value per grouping row by construction, so there is nothing for a dialect to
convert. A non-scalar summary, and any other runtime-only incompatibility,
therefore still surfaces from the database when the staged query executes.

The portable value contract covers finite numbers, missing values, and zero
denominators. Backend-specific non-finite representations are outside that
contract. See the
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
for both helpers' supported sources, denominator rules, `across()` forms,
empty inputs, and post-summary `mutate()` guidance.

## Completion stays in the lazy input pipeline

marginplyr does not create absent dimension keys or move completion data
between R and a database. Use a same-connection dimension relation, or move a
local relation explicitly before the Margin operation. The [Complete absent
keys before margins
guide](https://sayuks.github.io/marginplyr/vignettes/completing_keys.html)
shows a lazy dimension-table union and join, `dbplyr::copy_inline()` for small
local keys, and explicit `dplyr::copy_to()` for larger or reusable keys.

## Other lazy backends

dtplyr is lazy without being a SQL connection. `show_query()` displays the
generated data.table expression:

```{r}
#| eval: !expr has_dtplyr

dt_query <- january_sales |>
  dtplyr::lazy_dt() |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

show_query(dt_query)
collect(dt_query)
```

`nest_with_margins()` also supports dtplyr and remains lazy until collection.
SQL database and Arrow tables cannot use the nesting verbs because their
result contains a list column, and the refusal names the classes that can:

```{r}
#| label: nesting-a-sql-table
#| must_error: marginplyr_error

postgres_sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )
```

dtplyr also enforces the contextual-share source rules, for both denominators.
A valid request stays a native lazy query, and an ineligible source type or a
non-scalar summary raises a Package condition at explicit execution, before an
invalid grouping row is emitted:

```{r}
#| label: dtplyr-ineligible-share-source
#| eval: !expr has_dtplyr
#| must_error: marginplyr_error

january_sales |>
  dtplyr::lazy_dt() |>
  summarize_with_margins(
    last_channel = max(channel),
    last_channel_share = share_of_parent(last_channel),
    .grouping = rollup(region, store)
  ) |>
  collect()
```

Arrow summaries and expansions are supported and lazy wherever Arrow can
evaluate the summary itself. Two things are refused instead, for two unrelated
reasons.

The first is a summary Arrow's own engine cannot evaluate. Arrow answers one of
those by reading the whole input — every column, not just the ones the summary
names — and computing it in R, so marginplyr refuses it before a row is read
and tells you the two ways to get it computed:

```{r}
#| label: arrow-absorbed-summary
#| eval: !expr has_arrow
#| must_error: marginplyr_error

arrow::arrow_table(january_sales) |>
  summarize_with_margins(
    channels = paste(sort(unique(channel)), collapse = "/"),
    .grouping = rollup(region, store)
  )
```

Collecting first is one rewrite; selecting the columns the summary needs before
collecting is the other, and it is the one Arrow's own route cannot take for
you. Which expressions this reaches is Arrow's to decide and changes with its
version, so the shapes are described in one place rather than repeated here —
see the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html).
An Arrow dataset answers the same expressions with Arrow's own error rather
than this one, because a dataset never reads itself into R.

The second is contextual shares. `share_of_parent()` and `share_of_total()` are
both rejected outright, before any Arrow query is constructed: the reason is the
source summary they share. Collect the Arrow table first when you need either.

```{r}
#| label: arrow-parent-share
#| eval: !expr has_arrow
#| must_error: marginplyr_error

arrow::arrow_table(january_sales) |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_share = share_of_parent(revenue),
    .grouping = rollup(region, store)
  )
```

```{r}
#| label: arrow-total-share
#| eval: !expr has_arrow
#| must_error: marginplyr_error

arrow::arrow_table(january_sales) |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    revenue_of_total = share_of_total(revenue),
    .grouping = rollup(region, store)
  )
```

The
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
states where each backend reports these errors.

## Backend verification levels

Support is not one status. SQL simulation is useful evidence about rendering,
but it is not the same as running a query on a live server, so it is worth
knowing which claim rests on which.

Only DuckDB and SQLite are exercised as live SQL databases. DuckDB covers
native SQL end to end — generation, execution, result values, labels, and
grouping identity — while SQLite's live claim is narrower: portable
`UNION ALL` execution for Parent shares, including missing-safe matching and
Grouping set identifiers. PostgreSQL and the dozen-plus fallback dialects are
verified through dbplyr simulators, which prove what SQL is rendered and
nothing about a running server. Arrow and dtplyr are tested for the lazy
behavior each supports.

Treat a simulator-only backend as SQL-generation support, and report
server-specific behavior with a minimal query and the backend version.
*Database backend coverage* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
names every verified backend.

## Practical rules for remote analysis

- Build the complete lazy pipeline before calling `collect()`.

- Use `show_query()` to confirm native `GROUPING SETS` or fallback
  `UNION ALL`.

- Use `.by` for columns that remain fixed in every grouping level.

- Keep `grouping_bit()` or `grouping_id()` when source missing values or the
  literal label `"Total"` can occur in the data.

- Lazy results have no guaranteed order unless the query explicitly requests
  one. Ask for a Margin order with `.sort`, or call `arrange()` for any other
  presentation order.

- Lazy inputs default to `.check_margin_label = FALSE`, which skips only the
  *observed* half of the collision check; a label equal to a *declared*
  factor level is rejected either way. See *`.check_margin_label` on lazy
  inputs* above.

- A contextual share needs `.check_share_source = FALSE` on a backend that
  cannot establish that its source is eligible: a connection that executes
  nothing, such as a `simulate_*()` one, and a dialect that converts an
  ineligible source rather than refusing it, such as SQLite. See *Parent
  shares use a staged lazy query* above.

- A fallback query may repeat the source relation once per grouping set.
  Inspect the generated SQL and the database query plan for large workloads;
  marginplyr does not claim that one translation is universally faster.
