---
title: Database and lazy backends
vignette: >
  %\VignetteIndexEntry{Database and lazy backends}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` applies one grouping plan to local data frames and supported lazy
tables. With a database table, the plan remains lazy: `show_query()` renders
the SQL, while `collect()` executes it and brings the result into R.

If grouping sets are new to you, read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
first. This guide focuses on query shape, execution, and backend support.

## Installation

Install the development version of marginplyr from GitHub:

```{r}
#| eval: false

install.packages("pak")
pak::pkg_install("sayuks/marginplyr")

# Used for the live DuckDB example in this guide.
install.packages(c("DBI", "duckdb"))
```

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

## One report, two execution locations

The local report uses store detail, region subtotals, and a company total:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

local_result <- january_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store),
    .sort = TRUE
  )

local_result
```

The grouping expression does not change when the input becomes a lazy table:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_postgres()
)

postgres_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store),
    .sort = TRUE
  )

postgres_query
```

Printing a lazy table may show a preview, depending on the backend.
`show_query()` is the deliberate way to inspect the query without collecting
the result:

```{r}
show_query(postgres_query)
```

Use `show_query()` during interactive analysis. Use
`dbplyr::sql_render()` when code needs the rendered SQL value for logging,
testing, or another programmatic purpose.

## Native grouping sets

DuckDB and PostgreSQL use marginplyr's native path. The generated query
contains one `GROUP BY GROUPING SETS` operation. The requested rollup:

```text
(region, store)
(region)
()
```

becomes the equivalent SQL grouping-set list. Hidden `GROUPING()` flags allow
the outer query to label only dimensions removed by aggregation. A source
`NULL` therefore stays different from a generated subtotal even though SQL
would otherwise display both as missing.

More complex SQL grouping expressions use the same translation. The two
rollups below are combined by Cartesian product, matching comma-separated SQL
`GROUP BY` items:

```{r}
dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
) |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  ) |>
  show_query()
```

`grouping_sets(rollup(...), rollup(...))` has different semantics: it forms
the union of two independent hierarchies. See [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html#union-versus-cartesian-product)
for the complete comparison.

## Execute in DuckDB

The simulator above verifies SQL rendering without a server. DuckDB lets the
same workflow execute against a live in-memory database:

```{r}
con <- DBI::dbConnect(duckdb::duckdb())

sales_db <- copy_to(
  con,
  retail_sales,
  name = "retail_sales",
  temporary = TRUE,
  overwrite = TRUE
)

duckdb_query <- sales_db |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store),
    .sort = TRUE
  )

show_query(duckdb_query)
```

At this point `duckdb_query` is still lazy. `collect()` asks DuckDB to execute
the grouping sets and transfers only the summarized result:

```{r}
duckdb_result <- collect(duckdb_query)
DBI::dbDisconnect(con)

duckdb_result
```

Use `compute()` instead when the next stage should remain in the database as a
materialized or temporary remote table.

## Portable fallback SQL

marginplyr does not assume that every SQL dialect supports native grouping
sets. For an unconfirmed backend, it expands the plan into one query branch
per grouping set and joins those branches with `UNION ALL`.

```{r}
sqlite_sales <- dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
)

sqlite_query <- sqlite_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )

show_query(sqlite_query)
```

This SQL is longer than the native form, but it preserves the same requested
grouping levels and margin labels. Native versus fallback translation is
selected by the backend adapter; analysis code does not need a different
grouping specification.

`expand_with_margins()` also stays lazy. On a fallback backend, its SQL makes
the row-expansion branches explicit:

```{r}
sqlite_sales |>
  expand_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

## Other lazy backends

dtplyr is lazy without being a SQL connection. `show_query()` displays the
generated data.table expression:

```{r}
dt_query <- january_sales |>
  dtplyr::lazy_dt() |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

show_query(dt_query)
collect(dt_query)
```

`nest_with_margins()` also supports dtplyr and remains lazy until collection.
SQL database and Arrow tables cannot use the nesting verbs because their
result contains a list column.

## Backend verification levels

The status describes what the automated test suite actually verifies. SQL
simulation is useful evidence about rendering, but it is not the same as
running a query on a live server.

| Status | Backends | What is verified |
|---|---|---|
| Live native execution tested | DuckDB | SQL generation, execution, result values, labels, and grouping identity |
| Native SQL generation tested | PostgreSQL | SQL rendered with dbplyr's PostgreSQL simulator |
| Fallback SQL generation tested | Access, SAP HANA, Hive, Impala, MariaDB, Microsoft SQL Server, MySQL, Oracle, Amazon Redshift, Snowflake, Spark SQL, SQLite, Teradata, and generic DBI/ODBC connections | Portable branch and `UNION ALL` rendering with dbplyr simulators |
| Non-SQL lazy backend tested | Arrow, dtplyr | Lazy summary or expansion behavior appropriate to the backend |

Only DuckDB is currently exercised as a live SQL database in automated tests.
Treat simulator-only backends as SQL-generation support, and report
server-specific behavior with a minimal query and the backend version.

## Practical rules for remote analysis

- Build the complete lazy pipeline before calling `collect()`.

- Use `show_query()` to confirm native `GROUPING SETS` or fallback
  `UNION ALL`.

- Use `.by` for columns that remain fixed in every grouping level.

- Keep `grouping_bit()` or `grouping_id()` when source missing values or the
  literal label `"Total"` can occur in the data.

- Lazy results have no guaranteed order unless the query explicitly requests
  one. Set `.sort = TRUE` or call `arrange()` when presentation order matters.

- Lazy inputs default to `.check_margin_label = FALSE` because detecting an
  existing label would require an additional query. Opt in only when that
  extra scan is appropriate.

- A fallback query may repeat the source relation once per grouping set.
  Inspect the generated SQL and the database query plan for large workloads;
  marginplyr does not claim that one translation is universally faster.
