---
title: Get started
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` brings SQL-style grouping sets to `dplyr` workflows. A single
grouping specification can be executed locally, translated to native SQL, or
expanded to a portable `UNION ALL` query.

```{r setup}
library(marginplyr)
library(dplyr)
library(tibble)
```

```{r}
sales <- tribble(
  ~year, ~region, ~store, ~product, ~channel, ~amount,
  2024L, "East", "E1", "A", "Online", 10,
  2024L, "East", "E2", "B", "Store", 20,
  2024L, "West", "W1", "A", "Online", 15,
  2025L, "East", "E1", "A", "Store", 12,
  2025L, "West", "W1", NA_character_, "Online", 18
)
```

## The interface

`.by` contains fixed groups that occur in every result level. `.grouping`
describes the dimensions that can be aggregated.

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .by = year,
    .grouping = rollup(region, store)
  )
```

`"Total"` is a display label. Internally, grouping rows are identified from
the grouping plan or SQL `GROUPING()` result, never by comparing this string.

## Arbitrary grouping sets

`grouping_sets()` forms the union of explicitly requested sets.
`grouping_set()` with no columns is the empty grouping set and produces a
grand total (or one total per fixed `.by` group).

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = grouping_sets(
      grouping_set(region, store),
      grouping_set(region, product),
      grouping_set()
    )
  )
```

## ROLLUP

`rollup(region, store)` expands to the following sets:

```text
(region, store)
(region)
()
```

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = rollup(region, store)
  )
```

## CUBE

`cube(product, channel)` expands to every subset:

```text
(product, channel)
(product)
(channel)
()
```

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = cube(product, channel)
  )
```

## Cartesian products

SQL combines comma-separated grouping items by Cartesian product.
`grouping_spec()` provides the same operation in R.

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .by = year,
    .grouping = grouping_spec(
      rollup(region, store),
      cube(product, channel)
    )
  )
```

The compiler expands the product to concrete grouping sets before choosing a
local or SQL execution adapter.

## Nested specifications

Grouping families can be nested in `grouping_sets()`.

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = grouping_sets(
      rollup(region, store),
      cube(product, channel)
    ),
    .duplicates = "drop"
  )
```

Both families contain the empty set. The default `.duplicates = "error"`
detects this; `"drop"` keeps the first occurrence and `"keep"` preserves every
occurrence.

## Composite dimensions

A nested `grouping_set()` inside `rollup()` or `cube()` is one indivisible
dimension.

```{r}
locations <- tibble(
  country = c("JP", "JP", "US"),
  state = c("Tokyo", "Osaka", "CA"),
  year = c(2024L, 2025L, 2024L),
  amount = c(10, 20, 30)
)

locations |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = cube(grouping_set(country, state), year)
  )
```

`country` and `state` are always included or aggregated together.

## Inspecting aggregation levels

`grouping(x)` returns 1 when `x` was aggregated away and 0 when it belongs to
the active grouping set. `grouping_id()` combines those flags as a bit mask;
the last argument is the least-significant bit.

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    region_grouped = grouping(region),
    level = grouping_id(region, store, product),
    .grouping = grouping_spec(
      rollup(region, store),
      cube(product)
    )
  )
```

This also distinguishes source missing values from grouping rows. A source
`NA` has a grouping bit of 0; an `NA` produced by aggregation has a bit of 1.

## Labels and column types

Change the report label with `.margin_label`:

```{r}
sales |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = rollup(region, store),
    .margin_label = "All"
  )
```

When a character label is used, grouping dimensions are returned as character
columns; local factors are reconstructed with the label as their first level.
Use `.margin_label = NULL` to keep original column types and typed missing
values. In that mode, use `grouping()` or `grouping_id()` to tell source and
aggregation missing values apart.

For local data, `.check_margin_label = TRUE` prevents an existing value such
as `"Total"` from becoming visually ambiguous. Disable the check only when the
grouping bits are also retained.

## Computed grouping columns

Grouping constructors accept column selections rather than arbitrary SQL
expressions. Create an expression result first with `mutate()`.

```{r}
sales |>
  mutate(region_product = paste(region, product, sep = ":")) |>
  summarize_with_margins(
    total = sum(amount),
    .grouping = rollup(region_product)
  )
```

This form also translates predictably for lazy database tables.

## Other verbs

The same plan is shared by all margin verbs.

```{r}
sales |>
  union_all_with_margins(
    .grouping = rollup(region, store)
  )

sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

sales |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  )
```

`union_all_with_margins()` emits one input-row copy per grouping set.
`nest_with_margins()` returns an ordinary nested result, while
`nest_by_with_margins()` returns its row-wise counterpart.

## Database execution

DuckDB and PostgreSQL currently use native `GROUP BY GROUPING SETS`. Native
queries use hidden `GROUPING()` flags and apply the display label in an outer
query, preserving the input type seen by summary expressions. Other or
unconfirmed SQL dialects use the `UNION ALL` adapter. This conservative
capability policy favors equal results over assuming dialect support.
