---
title: Get started
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` brings SQL-style `GROUPING SETS`, `ROLLUP`, and `CUBE` to
`dplyr`. This guide follows one retail reporting task from a local data frame
to generated SQL.

## Installation

Install the development version from GitHub:

```{r}
#| eval: false

# install.packages("pak")
pak::pak("sayuks/marginplyr")

# Optional packages used in the backend and reporting sections
install.packages(c("DBI", "duckdb", "dtplyr", "knitr", "quartabs"))
```

```{r setup}
library(marginplyr)
library(dplyr)
```

## The sales report

`retail_sales` contains two years of monthly sales for a fictional retail
chain. Each row records one product, channel, and store combination.

```{r}
retail_sales
```

Online-direct sales do not belong to a physical store, so some source rows
have `store = NA`. This intentional missing value becomes important when
identifying subtotal rows.

### One independent report per month

The operations team first wants store detail, region subtotals, and a company
total for every month:

```{r}
monthly_report <- retail_sales |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .by = c(year, month),
    .grouping = rollup(region, store)
  )

monthly_report |>
  filter(year == 2026L, month == "Jan")
```

Columns in `.by` belong to every grouping set. They retain their input types
and never receive the margin label. Here, `rollup(region, store)` expands to:

```text
(region, store)
(region)
()
```

The empty set still includes the fixed `.by` columns, so it produces one
company total per year and month.

### Extend the hierarchy through time

The annual review also needs monthly, annual, and all-period totals. Moving
the period columns from `.by` into `rollup()` creates one longer hierarchy:

```{r}
all_levels <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    level = grouping_id(year, month, region, store),
    .grouping = rollup(year, month, region, store)
  )

all_levels |>
  filter(month == "Total")
```

The rollup expands to:

```text
(year, month, region, store)  store detail
(year, month, region)         monthly region subtotal
(year, month)                 monthly company total
(year)                        annual company total
()                            all-period company total
```

Argument order defines the hierarchy. This rollup does not create an annual
total for each region; arbitrary grouping sets can request that separately.

## Choose the grouping operation

### Arbitrary grouping sets

Finance needs monthly company totals, region-product totals, and the overall
grand total—but no other combinations. `grouping_sets()` forms the union of
exactly those requests:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      grouping_set(year, month),
      grouping_set(region, product),
      grouping_set()
    )
  )
```

An empty `grouping_set()` represents `()` and produces the grand total.

### CUBE

Merchandising wants every subtotal across product and sales channel.
`cube(product, channel)` expands to all four subsets:

```text
(product, channel)
(product)
(channel)
()
```

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(product, channel)
  )
```

### Union versus Cartesian product

Two rollups inside `grouping_sets()` remain independent:

```{r}
independent_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "drop"
  )

independent_totals
```

This is a union of:

```text
(year, month), (year), (), (region, store), (region)
```

Both rollups contain `()`. The default `.duplicates = "error"` catches this
overlap. `"drop"` retains one copy; `"keep"` retains both copies when repeated
totals are intentionally meaningful:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "keep"
  ) |>
  filter(
    year == "Total",
    region == "Total"
  )
```

SQL combines comma-separated grouping items by Cartesian product.
`grouping_spec()` provides that operation:

```{r}
combined_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  )

combined_totals |>
  slice_head(n = 12)
```

The two families each contain three sets, so their Cartesian product contains
nine sets:

```text
(year, month, region, store)
(year, month, region)
(year, month)
(year, region, store)
(year, region)
(year)
(region, store)
(region)
()
```

### Composite dimensions and tidy-select

A `grouping_set()` nested directly inside `rollup()` or `cube()` is one
indivisible dimension. The following cube includes or removes `region` and
`store` together:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(
      grouping_set(region, store),
      all_of(c("product", "channel"))
    )
  ) |>
  slice_head(n = 12)
```

Grouping specifications accept tidy-select expressions. They select existing
columns; they do not evaluate arbitrary SQL expressions.

### Computed grouping columns

Create a computed dimension with `mutate()` before summarizing:

```{r}
retail_sales |>
  mutate(period = paste(year, month, sep = "-")) |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(period, region)
  )
```

This two-step form works consistently for local and lazy tables.

## Identify totals without guessing from labels

`grouping(x)` returns `1` when `x` was removed from the current grouping set
and `0` when it remains part of the group. It does not test whether the
visible value is missing.

```{r}
retail_sales |>
  filter(
    year == 2026L,
    month == "Jan",
    region == "East"
  ) |>
  summarize_with_margins(
    revenue = sum(revenue),
    store_is_total = grouping(store),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )
```

The source `NA` store has `store_is_total = 0`. The region subtotal has
`store_is_total = 1`, because `store` was removed by the rollup.

`grouping_id()` combines the flags as a bit mask. Its last argument is the
least-significant bit:

```text
region  store  bits  grouping_id
kept    kept    00       0
kept    total   01       1
total   total   11       3
```

Columns fixed in `.by` can also be passed to these helpers. Their flag is
always zero because they occur in every grouping set.

### Labels and column types

Change the label used for dimensions removed by aggregation:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(year),
    .margin_label = "All years"
  )
```

Use `.margin_label = NULL` to preserve the grouping column's input type:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    year_is_total = grouping(year),
    .grouping = rollup(year),
    .margin_label = NULL
  )
```

With a display label, grouping dimensions become character columns. With
`NULL`, both source and generated missing values remain typed missing values,
so retain `grouping()` or `grouping_id()` when the distinction matters.

For local data, `.check_margin_label = TRUE` prevents an existing value such
as `"Total"` from becoming visually ambiguous. `.sort` controls the optional
local result ordering; lazy tables remain unordered unless explicitly
arranged.

## Inspect the generated SQL

### PostgreSQL without a server

dbplyr's simulated backends show generated SQL without a live database. A
PostgreSQL simulation uses marginplyr's native translation:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  )

show_query(postgres_query)
```

The inner query contains one `GROUP BY GROUPING SETS` clause. Hidden
`GROUPING()` flags let the outer query label only generated subtotal values.

### A portable fallback

For dialects without confirmed native support, marginplyr emits one branch per
grouping set and combines them with `UNION ALL`:

```{r}
sqlite_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_sqlite()
)

sqlite_query <- sqlite_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = rollup(region, store)
  )

show_query(sqlite_query)
```

This conservative fallback favors consistent result semantics over assuming
that an unconfirmed backend implements grouping sets.

### Execute the query in DuckDB

DuckDB supports native grouping sets. The following example copies the small
data set into an in-memory database, displays the SQL, and executes it:

```{r}
local({
  con <- suppressMessages(DBI::dbConnect(duckdb::duckdb()))
  on.exit(DBI::dbDisconnect(con), add = TRUE)

  sales_db <- copy_to(
    con,
    retail_sales,
    name = "retail_sales",
    temporary = TRUE,
    overwrite = TRUE
  )

  duckdb_query <- sales_db |>
    filter(year == 2026L, month == "Jan") |>
    summarize_with_margins(
      revenue = sum(revenue, na.rm = TRUE),
      level = grouping_id(region, store),
      .grouping = rollup(region, store),
      .sort = TRUE
    )

  show_query(duckdb_query)
  collect(duckdb_query)
})
```

The result stays lazy until `collect()` or another execution verb is called.

## Keep the detail rows

### Expand rows with `union_all_with_margins()`

`union_all_with_margins()` emits one copy of each source row for every
grouping set. It is useful when downstream work needs the rows rather than an
immediate summary:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

expanded <- january_sales |>
  union_all_with_margins(
    .grouping = rollup(region, store)
  )

expanded |>
  count(region, store, name = "records")
```

The same verb stays lazy for database tables. Its SQLite SQL makes the
row-expansion branches explicit:

```{r}
dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
) |>
  union_all_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

### Nest locally or with dtplyr

`nest_with_margins()` keeps one detail table for each visible group:

```{r}
january_sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )
```

A dtplyr input performs the expansion and nesting lazily until `collect()`:

```{r}
january_sales |>
  dtplyr::lazy_dt() |>
  nest_with_margins(
    .grouping = rollup(region, store)
  ) |>
  collect()
```

`nest_by_with_margins()` returns a row-wise data frame, making one calculation
per nested report section straightforward:

```{r}
january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  ) |>
  mutate(
    records = nrow(data),
    units = sum(data$units),
    revenue = sum(data$revenue)
  ) |>
  select(-data)
```

The row-wise result is local even when the input is a dtplyr step:

```{r}
january_sales |>
  dtplyr::lazy_dt() |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  ) |>
  mutate(
    records = nrow(data),
    revenue = sum(data$revenue)
  ) |>
  select(-data)
```

List columns are why the two nesting verbs support local data frames and
dtplyr steps rather than SQL database tables.

## Optional Quarto tabsets with quartabs

[`quartabs`](https://sayuks.github.io/quartabs/) can turn the nested report
sections into Quarto tabs. This compact example creates one tab per region
plus a company-total tab.

The chunk option `results: asis` is required because `render_tabset()` emits
Quarto tabset markup.

```{r}
#| results: asis

tabbed_report <- january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region)
  ) |>
  mutate(
    region = factor(region, levels = c("Total", "East", "West")),
    report = list(
      data |>
        summarize(
          records = n(),
          units = sum(units),
          revenue = sum(revenue)
        ) |>
        knitr::kable()
    )
  ) |>
  ungroup()

tabbed_report |>
  quartabs::render_tabset(
    tabset_vars = region,
    output_vars = report
  )
```

The `"Total"` tab is created by the grouping plan. The ordered factor makes it
the first tab; `quartabs` only handles presentation, while marginplyr defines
which source rows belong to that total.
