---
title: Get started
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

`marginplyr` brings SQL-style `GROUPING SETS`, `ROLLUP`, and `CUBE` to
`dplyr`. This guide follows one retail reporting task from a local data frame
to generated SQL.

## Installation

Install the development version from GitHub:

```{r}
#| eval: false

if (!"remotes" %in% rownames(installed.packages())) {
  install.packages("remotes")
}
remotes::install_github("sayuks/marginplyr")

# Install only packages that are missing from the backend and reporting
# sections below.
guide_packages <- c("DBI", "duckdb", "dtplyr", "knitr", "quartabs")
missing_packages <- setdiff(guide_packages, rownames(installed.packages()))
if (length(missing_packages) > 0L) {
  install.packages(missing_packages)
}
```

```{r setup}
library(marginplyr)
library(dplyr)
```

## The sales report

`retail_sales` contains two years of monthly sales for a fictional retail
chain. Each row records one product, channel, and store combination.

```{r}
retail_sales
```

Online-direct sales do not belong to a physical store, so some source rows
have `store = NA`. This intentional missing value becomes important when
identifying subtotal rows.

### One independent report per month

The operations team first wants store detail, region subtotals, and a company
total for every month:

```{r}
monthly_report <- retail_sales |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .by = c(year, month),
    .grouping = rollup(region, store)
  )

monthly_report |>
  filter(year == 2026L, month == "Jan")
```

Columns in `.by` belong to every grouping set. They retain their input types
and never receive the margin label. Here, `rollup(region, store)` expands to:

```text
(region, store)
(region)
()
```

The empty set still includes the fixed `.by` columns, so it produces one
company total per year and month.

### Persistent groups as fixed keys

Existing `dplyr::group_by()` columns provide the same fixed keys when `.by` is
not supplied. This is useful when a pipeline has already established its
report partitions:

```{r}
grouped_monthly_report <- retail_sales |>
  group_by(year, month) |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

grouped_monthly_report |>
  filter(year == 2026L, month == "Jan")

group_vars(grouped_monthly_report)
```

The calculation is equivalent to the earlier `.by = c(year, month)` call.
Do not combine a grouped input with `.by`; this is an error, matching
`dplyr::summarize()` and `tidyr::nest()`. A fixed grouping column also cannot
appear in `.grouping`; call `ungroup()` first if that column should be rolled
up instead. Row-wise inputs must likewise be ungrouped explicitly. Local
groups created with `.drop = FALSE` are rejected because empty factor groups
do not have a consistent SQL or lazy-backend equivalent.

Input groups affect the calculation but, except for the explicitly row-wise
nesting verb, do not persist on the result:

| Function | Output grouping | Difference from the closest official verb |
|---|---|---|
| `summarize_with_margins()` | Ungrouped | `dplyr::summarize()` can retain groups through `.groups`; arbitrary grouping sets have no single `drop_last` level |
| `expand_with_margins()` | Ungrouped | The expanded branches deliberately discard persistent grouping metadata |
| `nest_with_margins()` | Ungrouped | `tidyr::nest()` preserves grouped input; this function consumes those groups as fixed keys but does not preserve them |
| `nest_by_with_margins()` | Row-wise by every visible key | Like `dplyr::nest_by()`, but grouping-set dimensions are also outer row-wise keys |

This invariant is the same for local data frames, dtplyr, Arrow, and SQL lazy
tables. The two nesting functions support local data frames and dtplyr because
SQL and Arrow tables cannot contain their list-column result.

### Intentional differences from `dplyr::summarize()`

The summary expressions still use dplyr data masking, but marginplyr narrows a
few parts of the official API where arbitrary grouping sets do not have one
portable interpretation:

| Feature | Official dplyr behavior | marginplyr behavior and rationale |
|---|---|---|
| Spelling | `summarise()` and `summarize()` are synonyms | `summarise_with_margins()` and `summarize_with_margins()` are synonyms |
| `.groups` | Can return `"drop_last"`, `"drop"`, `"keep"`, or `"rowwise"` groups | Accepts only `NULL` or `"drop"` and always returns ungrouped output; multiple grains have no single grouping hierarchy |
| Summary name matches a grouping key | The local backend can overwrite the key | Rejected, including data-frame-valued summaries, so grouping-set identity is not destroyed and backend behavior remains portable |
| `across()` and `pick()` | Exclude the grouping columns active in the verb | Exclude all fixed keys and margin dimensions in the complete plan, even in a branch where a dimension is rolled up |
| `cur_group*()` and deprecated `cur_data*()` helpers | Describe the current dplyr group or data mask | Rejected because branch-local identifiers, rows, and columns have no global meaning after grouping sets are combined; use `grouping_bit()` or `grouping_id()` for margin levels |
| Ordering | `.by` summaries use first-appearance order | `.sort = FALSE` preserves first appearance locally; `.sort = TRUE` explicitly sorts; lazy order is otherwise unspecified |
| Backend extension | `summarize()` is a public S3 generic | The public verbs are regular functions; a private adapter handles schema and execution capabilities after one common plan is validated |

For example, `everything()` below selects measures but not `year`, `region`,
or `store`:

```{r}
retail_sales |>
  summarize_with_margins(
    across(
      everything(),
      n_distinct,
      .names = "distinct_{.col}"
    ),
    selected_measures = paste(
      names(pick(everything())),
      collapse = ", "
    ),
    .by = year,
    .grouping = rollup(region, store),
    .sort = FALSE
  ) |>
  slice_head(n = 6)
```

A summary may return a one-row data frame as in dplyr, but none of its output
columns may be named `year`, `region`, or `store` in this example. Use a new
summary name or rename the input key first. The private adapter is
deliberately not an extension API: adding a backend to marginplyr keeps
duplicate handling, labels, grouping flags, and output grouping under the same
public contract.

### Extend the hierarchy through time

The annual review also needs monthly, annual, and all-period totals. Moving
the period columns from `.by` into `rollup()` creates one longer hierarchy:

```{r}
all_levels <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    level = grouping_id(year, month, region, store),
    .grouping = rollup(year, month, region, store)
  )

all_levels |>
  filter(month == "Total")
```

The rollup expands to:

```text
(year, month, region, store)  store detail
(year, month, region)         monthly region subtotal
(year, month)                 monthly company total
(year)                        annual company total
()                            all-period company total
```

Argument order defines the hierarchy. This rollup does not create an annual
total for each region; arbitrary grouping sets can request that separately.

## Choose the grouping operation

### Arbitrary grouping sets

Finance needs monthly company totals, region-product totals, and the overall
grand total—but no other combinations. `grouping_sets()` forms the union of
exactly those requests:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      grouping_set(year, month),
      grouping_set(region, product),
      grouping_set()
    )
  )
```

An empty `grouping_set()` represents `()` and produces the grand total.

### CUBE

Merchandising wants every subtotal across product and sales channel.
`cube(product, channel)` expands to all four subsets:

```text
(product, channel)
(product)
(channel)
()
```

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(product, channel)
  )
```

### Union versus Cartesian product

Two rollups inside `grouping_sets()` remain independent:

```{r}
independent_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "drop"
  )

independent_totals
```

This is a union of:

```text
(year, month), (year), (), (region, store), (region)
```

Both rollups contain `()`. The default `.duplicates = "error"` catches this
overlap. `"drop"` retains one copy; `"keep"` retains both copies when repeated
totals are intentionally meaningful:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "keep"
  ) |>
  filter(
    year == "Total",
    region == "Total"
  )
```

SQL combines comma-separated grouping items by Cartesian product.
`grouping_spec()` provides that operation:

```{r}
combined_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  )

combined_totals |>
  slice_head(n = 12)
```

The two families each contain three sets, so their Cartesian product contains
nine sets:

```text
(year, month, region, store)
(year, month, region)
(year, month)
(year, region, store)
(year, region)
(year)
(region, store)
(region)
()
```

### Composite dimensions and tidy-select

A `grouping_set()` nested directly inside `rollup()` or `cube()` is one
indivisible dimension. The following cube includes or removes `region` and
`store` together:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(
      grouping_set(region, store),
      all_of(c("product", "channel"))
    )
  ) |>
  slice_head(n = 12)
```

Grouping specifications accept tidy-select expressions. They select existing
columns; they do not evaluate arbitrary SQL expressions.

### Computed grouping columns

Create a computed dimension with `mutate()` before summarizing:

```{r}
retail_sales |>
  mutate(period = paste(year, month, sep = "-")) |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(period, region)
  )
```

This two-step form works consistently for local and lazy tables.

## Identify totals without guessing from labels

`grouping_bit(x)` corresponds to SQL `GROUPING(x)`: it returns `1` when `x`
was removed from the current grouping set and `0` when it remains part of the
group. It does not test whether the visible value is missing.

The `_bit` suffix describes this `0`/`1` result and avoids masking the
unrelated `base::grouping()` function, which returns a permutation for placing
equal values next to one another.

```{r}
retail_sales |>
  filter(
    year == 2026L,
    month == "Jan",
    region == "East"
  ) |>
  summarize_with_margins(
    revenue = sum(revenue),
    store_is_total = grouping_bit(store),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )
```

The source `NA` store has `store_is_total = 0`. The region subtotal has
`store_is_total = 1`, because `store` was removed by the rollup.

`grouping_id()` combines the flags as a bit mask. Its last argument is the
least-significant bit:

```text
region  store  bits  grouping_id
kept    kept    00       0
kept    total   01       1
total   total   11       3
```

Columns fixed in `.by` can also be passed to these helpers. Their flag is
always zero because they occur in every grouping set.

### Labels and column types

Change the label used for dimensions removed by aggregation:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(year),
    .margin_label = "All years"
  )
```

Use `.margin_label = NULL` to preserve the grouping column's input type:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    year_is_total = grouping_bit(year),
    .grouping = rollup(year),
    .margin_label = NULL
  )
```

With a display label, grouping dimensions become character columns. With
`NULL`, both source and generated missing values remain typed missing values,
so retain `grouping_bit()` or `grouping_id()` when the distinction matters.

For local data, `.check_margin_label = TRUE` prevents an existing value such
as `"Total"` from becoming visually ambiguous. `.sort` controls the optional
result ordering. Lazy inputs default to `.check_margin_label = FALSE` because
checking would execute an additional query, and they remain unordered unless
explicitly arranged. A label is only a display value; retain `grouping_bit()`
or `grouping_id()` whenever source values can equal the label.

## Backend support

marginplyr uses two SQL execution paths:

| Execution path | Backends | Automated verification |
|---|---|---|
| Native `GROUP BY GROUPING SETS` | DuckDB, PostgreSQL | Live DuckDB execution; PostgreSQL SQL simulation |
| Portable `UNION ALL` | Access, SAP HANA, Hive, Impala, MariaDB, Microsoft SQL Server, MySQL, Oracle, Amazon Redshift, Snowflake, Spark SQL, SQLite, Teradata, and generic DBI/ODBC connections | SQL generation with dbplyr simulators |

Arrow and dtplyr are also tested lazy backends, but they are not SQL database
connections. SQL simulation confirms that marginplyr can render each listed
dialect; only DuckDB is currently exercised against a live database in the
automated test suite.

The public verbs are intentionally not backend S3 generics. Each call first
compiles one backend-independent grouping plan. A non-exported adapter then
chooses schema handling and either native grouping sets or the portable
branch expansion. This makes adding a backend a package-level compatibility
decision rather than allowing third-party methods to silently redefine label,
duplicate, or grouping semantics.

## Inspect the generated SQL

### PostgreSQL without a server

dbplyr's simulated backends show generated SQL without a live database. A
PostgreSQL simulation uses marginplyr's native translation:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_query <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  )

show_query(postgres_query)
```

The inner query contains one `GROUP BY GROUPING SETS` clause. Hidden
`GROUPING()` flags let the outer query label only generated subtotal values.

### A portable fallback

For dialects without confirmed native support, marginplyr emits one branch per
grouping set and combines them with `UNION ALL`:

```{r}
sqlite_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_sqlite()
)

sqlite_query <- sqlite_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = rollup(region, store)
  )

show_query(sqlite_query)
```

This conservative fallback favors consistent result semantics over assuming
that an unconfirmed backend implements grouping sets.

### Execute the query in DuckDB

DuckDB supports native grouping sets. The following example copies the small
data set into an in-memory database, displays the SQL, and executes it:

```{r}
con <- suppressMessages(DBI::dbConnect(duckdb::duckdb()))

sales_db <- copy_to(
  con,
  retail_sales,
  name = "retail_sales",
  temporary = TRUE,
  overwrite = TRUE
)

duckdb_query <- sales_db |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    level = grouping_id(region, store),
    .grouping = rollup(region, store),
    .sort = TRUE
  )

show_query(duckdb_query)
duckdb_result <- collect(duckdb_query)
DBI::dbDisconnect(con)
duckdb_result
```

The result stays lazy until `collect()` or another execution verb is called.

## Keep the detail rows

### Expand rows with `expand_with_margins()`

`expand_with_margins()` emits one copy of each source row for every grouping
set. It is useful when downstream work needs the rows rather than an
immediate summary. The public name describes the result; SQL `UNION ALL` is
only the portable implementation strategy and this function is unrelated to
`dplyr::union_all()`:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

expanded <- january_sales |>
  expand_with_margins(
    .grouping = rollup(region, store)
  )

expanded |>
  count(region, store, name = "records")
```

The same verb stays lazy for database tables. Its SQLite SQL makes the
row-expansion branches explicit:

```{r}
dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
) |>
  expand_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

### Nest locally or with dtplyr

`nest_with_margins()` keeps one detail table for each visible group:

```{r}
nested_sections <- january_sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

nested_sections |>
  mutate(records = vapply(data, nrow, integer(1))) |>
  select(-data)

# Inspect the source rows behind the first visible group.
nested_sections$data[[1]]
```

A dtplyr input performs the expansion and nesting lazily until `collect()`.
`show_query()` displays the generated data.table expression:

```{r}
nested_dt <- january_sales |>
  dtplyr::lazy_dt() |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

show_query(nested_dt)
collect(nested_dt)
```

`nest_by_with_margins()` returns a row-wise data frame, making one calculation
per nested report section straightforward:

```{r}
january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  ) |>
  mutate(
    records = nrow(data),
    units = sum(data$units),
    revenue = sum(data$revenue)
  ) |>
  select(-data)
```

The row-wise result is local even when the input is a dtplyr step:

```{r}
january_sales |>
  dtplyr::lazy_dt() |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  ) |>
  mutate(
    records = nrow(data),
    revenue = sum(data$revenue)
  ) |>
  select(-data)
```

Because `nest_by_with_margins()` collects before making the result row-wise,
its return value no longer has a lazy query to display. Use the
`nest_with_margins()` form immediately above with `show_query()` to inspect the
same dtplyr expansion and nesting expression before collection.

List columns are why the two nesting verbs support local data frames and
dtplyr steps rather than SQL database tables.

#### Relationship to tidyr and dplyr

The margin-aware verbs deliberately share the main result shapes of
`tidyr::nest()` and `dplyr::nest_by()`, but they are not drop-in replacements:

| Function | Official reference | Shared behavior | Deliberate differences |
|---|---|---|---|
| `nest_with_margins()` | `tidyr::nest()` | `NULL` `.key` becomes `"data"`; returns one list column | Uses `.by` plus `.grouping`; always nests all non-key columns; does not implement `...`, multiple list columns, `.names_sep`, or `.keep`; always returns ungrouped output |
| `nest_by_with_margins()` | `dplyr::nest_by()` | Requires a string `.key`; supports `.keep`; returns a row-wise data frame | Uses `.by` plus `.grouping` instead of data-masking `...`; collects a dtplyr result before making it row-wise |

The shared features in this table do not amount to complete upstream
compatibility; the deliberate differences are part of the margin-aware API.

For grouped input, existing grouping columns become fixed keys as though they
had been supplied through `.by`. Unlike `tidyr::nest()`, `nest_with_margins()`
does not preserve that grouping metadata on its output. Both nesting functions
sort keys by default through `.sort = TRUE`, whereas
`tidyr::nest(.by = ...)` preserves first-appearance order.

Only `nest_by_with_margins()` has `.keep`, matching `dplyr::nest_by()`. With
`.keep = TRUE`, nested tables retain the original pre-margin key values. For
example, the outer key of an `East / Total` subtotal displays `"Total"`, while
the rows inside contain their original store names. This preserves the
meaning of dplyr's “keep the grouping columns” rule after a dimension has been
rolled up.

Nesting accepts `.duplicates = "error"` or `"drop"`, but rejects `"keep"`.
Unlike a summary or row expansion, two duplicate nested groups have no
visible set identifier and would therefore be indistinguishable. Internal set
identifiers and `.keep` copies use collision-free temporary names, so no input
column name is reserved.

The list column is a regular list of data frames; a
`vctrs::list_of` subclass is not promised. On an empty ungrouped input,
`nest_with_margins()` follows `tidyr::nest()` and returns zero outer rows.
`nest_by_with_margins()` follows `dplyr::nest_by()` and returns one row
containing the empty data frame when there are no keys.

`dplyr::nest_by()` is currently not a stable upstream API and may eventually
be deprecated in favor of `tidyr::nest(.by = )`. marginplyr nevertheless
keeps its row-wise wrapper because per-margin modeling and reporting are
useful, and implements it through the same nesting core rather than a separate
semantic path.

## Optional Quarto tabsets with quartabs

[`quartabs`](https://sayuks.github.io/quartabs/) can turn the nested report
sections into Quarto tabs. This compact example creates one tab per region
plus a company-total tab.

When this guide is opened as an installed package vignette, Quarto's
lightweight vignette format does not provide interactive tabsets. The same
content is displayed sequentially instead. View the
[package-site version of this guide](https://sayuks.github.io/marginplyr/vignettes/get_started.html#optional-quarto-tabsets-with-quartabs)
to use the interactive tabs.

The chunk option `results: asis` is required because `render_tabset()` emits
Quarto tabset markup.

```{r}
#| results: asis

tabbed_report <- january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region)
  ) |>
  mutate(
    # Order the company total before regional tabs in the rendered report.
    region = factor(region, levels = c("Total", "East", "West")),
    report = list(
      data |>
        summarize(
          records = n(),
          units = sum(units),
          revenue = sum(revenue)
        ) |>
        knitr::kable()
    )
  ) |>
  ungroup()

tabbed_report |>
  quartabs::render_tabset(
    tabset_vars = region,
    output_vars = report
  )
```

The `"Total"` tab is created by the grouping plan. The ordered factor makes it
the first tab; `quartabs` only handles presentation, while marginplyr defines
which source rows belong to that total.
