---
title: Get started
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

Reports often need several grains at once: detail rows, subtotals, and a grand
total. `marginplyr` creates those levels with one dplyr-style summary. SQL
calls this idea `GROUPING SETS`; you do not need to know that syntax to begin.

This guide starts with a familiar reporting problem, then introduces
`rollup()`, `grouping_sets()`, and `cube()`. Database users can continue with
the [Database and lazy backends
guide](https://sayuks.github.io/marginplyr/vignettes/database_backends.html).

## Installation

Install the development version from GitHub:

```{r}
#| eval: false

install.packages("pak")
pak::pkg_install("sayuks/marginplyr")
```

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. A few sections below illustrate optional
tooling, so their chunks evaluate only when that tooling is installed. Where a
package is missing the code is still shown, but its output is absent.

```{r}
#| include: false

# dbplyr's SQLite simulator opens no connection, but rendering its SQL reads
# the installed driver version through RSQLite.
has_sqlite_simulator <- requireNamespace("RSQLite", quietly = TRUE)
has_dtplyr <- requireNamespace("dtplyr", quietly = TRUE)
has_tabsets <- requireNamespace("knitr", quietly = TRUE) &&
  requireNamespace("quartabs", quietly = TRUE)

# Chunks marked with `must_error` show a call that is meant to fail, so the
# reader sees the real diagnostic rather than a quotation of one. AGENTS.md is
# authoritative for what the option does and why it exists.
source(system.file("vignette-hooks", "must-error.R", package = "marginplyr"))
```

## Start with the report question

`retail_sales` contains two years of monthly sales for a fictional retail
chain. Each row records one product, channel, and store combination.

```{r}
retail_sales
```

Online-direct sales do not belong to a physical store, so some source rows
have `store = NA`. This intentional missing value becomes important later:
it is a real source value, not a subtotal.

### The manual version

For January 2026, the operations team wants store detail, region subtotals,
and a company total. With ordinary summaries, each level is a separate branch
that must be aligned and combined:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

manual_report <- bind_rows(
  january_sales |>
    summarize(revenue = sum(revenue), .by = c(region, store)),
  january_sales |>
    summarize(revenue = sum(revenue), .by = region) |>
    mutate(store = "Total") |>
    select(region, store, revenue),
  january_sales |>
    summarize(revenue = sum(revenue)) |>
    mutate(region = "Total", store = "Total") |>
    select(region, store, revenue)
)

manual_report
```

This is valid dplyr code and can be the clearest choice for one small,
one-off report. It becomes repetitive when the hierarchy grows, the requested
levels change, or every branch must stay lazy on a remote source.

### One grouping plan

`rollup(region, store)` describes the same three levels directly:

```{r}
january_report <- january_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

january_report
```

The result is one ordinary data frame. The grouping plan expands to:

```text
(region, store)  store detail
(region)         region subtotal
()               company total
```

The empty set `()` means that neither dimension remains, so it produces the
grand total.

### One independent report per month

The same hierarchy can be repeated for every reporting period without
building more branches:

```{r}
monthly_report <- retail_sales |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .by = c(year, month),
    .grouping = rollup(region, store)
  )

monthly_report |>
  filter(year == 2026L, month == "Jan")
```

Columns in `.by` belong to every grouping set. They retain their input types
and never receive the margin label. The empty set from `rollup()` still
includes the fixed `.by` columns, so it produces one company total per year
and month.

### The three grouping helpers

| Helper | Mental model | Typical question |
|---|---|---|
| `rollup()` | Ordered hierarchy | “Give me stores, regions, and the company.” |
| `grouping_sets()` | Exact union of requested levels | “Give me only these views.” |
| `cube()` | Every subset | “Give me every combination of these dimensions.” |

Most reporting tasks should start with `rollup()`. Reach for
`grouping_sets()` when a hierarchy includes too much or unrelated views must
share one result. Use `cube()` deliberately: the number of grouping sets
doubles with every independent dimension.

### Extend the hierarchy through time

The annual review also needs monthly, annual, and all-period totals. Moving
the period columns from `.by` into `rollup()` creates one longer hierarchy:

```{r}
all_levels <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    level = grouping_id(year, month, region, store),
    .grouping = rollup(year, month, region, store)
  )

all_levels |>
  filter(month == "Total")
```

The rollup expands to:

```text
(year, month, region, store)  store detail
(year, month, region)         monthly region subtotal
(year, month)                 monthly company total
(year)                        annual company total
()                            all-period company total
```

Argument order defines the hierarchy. This rollup does not create an annual
total for each region; arbitrary grouping sets can request that separately.

## Choose the grouping operation

### Arbitrary grouping sets

Finance needs monthly company totals, region-product totals, and the overall
grand total—but no other combinations. `grouping_sets()` forms the union of
exactly those requests:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      grouping_set(year, month),
      grouping_set(region, product),
      grouping_set()
    )
  )
```

An empty `grouping_set()` represents `()` and produces the grand total.

### CUBE

Merchandising wants every subtotal across product and sales channel.
`cube(product, channel)` expands to all four subsets:

```text
(product, channel)
(product)
(channel)
()
```

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(product, channel)
  )
```

### Union versus Cartesian product

Two rollups inside `grouping_sets()` remain independent:

```{r}
independent_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "drop"
  )

independent_totals
```

This is a union of:

```text
(year, month), (year), (), (region, store), (region)
```

Both rollups contain `()`. The default `.duplicates = "error"` catches this
overlap. `"drop"` retains one copy; `"keep"` retains both copies when repeated
totals are intentionally meaningful:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_sets(
      rollup(year, month),
      rollup(region, store)
    ),
    .duplicates = "keep"
  ) |>
  filter(
    year == "Total",
    region == "Total"
  )
```

SQL combines comma-separated grouping items by Cartesian product.
`grouping_spec()` provides that operation:

```{r}
combined_totals <- retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = grouping_spec(
      rollup(year, month),
      rollup(region, store)
    )
  )

combined_totals |>
  head(n = 12)
```

The two families each contain three sets, so their Cartesian product contains
nine sets:

```text
(year, month, region, store)
(year, month, region)
(year, month)
(year, region, store)
(year, region)
(year)
(region, store)
(region)
()
```

### Composite dimensions and tidy-select

A `grouping_set()` nested directly inside `rollup()` or `cube()` is one
indivisible dimension. The following cube includes or removes `region` and
`store` together:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(
      grouping_set(region, store),
      all_of(c("product", "channel"))
    )
  ) |>
  head(n = 12)
```

Grouping specifications accept tidy-select expressions. They select existing
columns; they do not evaluate arbitrary SQL expressions.

### Computed grouping columns

Create a computed dimension with `mutate()` before summarizing:

```{r}
retail_sales |>
  mutate(period = paste(year, month, sep = "-")) |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(period, region)
  )
```

This two-step form works consistently for local and lazy tables.

## When marginplyr fits

Several R workflows can produce subtotals. The useful distinction is not
whether they can calculate a total, but how the grouping levels are expressed
and where the calculation runs.

| Approach | Prefer it when | Tradeoff |
|---|---|---|
| Repeated `summarize()` plus `bind_rows()` | A one-off report has only a few levels and explicit branches are easiest to review | Every grouping level is authored and maintained separately |
| [`data.table`](https://rdatatable.gitlab.io/data.table/reference/groupingsets.html) | The analysis is a local data.table workflow | It uses the data.table API rather than a dbplyr lazy query |
| [`rollup`](https://cran.r-project.org/package=rollup) | Its list of grouped data frames is a natural intermediate representation | The workflow differs from marginplyr's single relational result |
| marginplyr | One grouping plan should work locally and through dbplyr, or grouping identity and margin-aware nesting matter | Its broader semantics add concepts that a single subtotal does not need |

Use the simplest approach that preserves the shape and execution model your
analysis needs. marginplyr is particularly useful when the report has many
grains, must stay remote until `collect()`, or needs to distinguish source
missing values from generated totals.

## Identify totals without guessing from labels

`grouping_bit(x)` corresponds to SQL `GROUPING(x)`: it returns `1` when `x`
was removed from the current grouping set and `0` when it remains part of the
group. It does not test whether the visible value is missing.

The `_bit` suffix describes this `0`/`1` result and avoids masking the
unrelated `base::grouping()` function, which returns a permutation for placing
equal values next to one another.

```{r}
retail_sales |>
  filter(
    year == 2026L,
    month == "Jan",
    region == "East"
  ) |>
  summarize_with_margins(
    revenue = sum(revenue),
    store_is_total = grouping_bit(store),
    level = grouping_id(region, store),
    .grouping = rollup(region, store)
  )
```

The source `NA` store has `store_is_total = 0`. The region subtotal has
`store_is_total = 1`, because `store` was removed by the rollup.

The [`grouping_bit()`
reference](https://sayuks.github.io/marginplyr/man/grouping_bit.html) compares
`.id`, `inspect_grouping()$set_id`, `grouping_bit()`, and `grouping_id()` in
full. The [Grouping identity
guide](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html)
works through the same values with executable examples, and
[Recipes](https://sayuks.github.io/marginplyr/vignettes/recipes.html) uses
them to keep one grouping set and to name the set behind every row.

`grouping_id()` combines the flags as a bit mask. Its last argument is the
least-significant bit:

```text
region  store  bits  grouping_id
kept    kept    00       0
kept    total   01       1
total   total   11       3
```

Columns fixed in `.by` can also be passed to these helpers. Their flag is
always zero because they occur in every grouping set.

### Labels and column types

Change the label used for dimensions removed by aggregation:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(year),
    .margin_label = "All years"
  )
```

Use a named vector to give each resolved Margin dimension its own label.
Names must cover the dimensions exactly once; `.by` columns are fixed keys
and cannot be named:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store),
    .margin_label = c(
      region = "All regions",
      store = "All stores"
    )
  )
```

For factors and ordered factors, a non-missing Margin label becomes a
synthetic level. It is placed last by default; use
`.margin_label_position = "first"` to put it first. This changes only the
level order, not the result row order.

Use `.margin_label = NULL` or `NA_character_` to preserve the grouping
column's input type and create a typed missing Margin value:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    year_is_total = grouping_bit(year),
    .grouping = rollup(year),
    .margin_label = NULL
  ) |>
  # `retail_sales` is a plain data frame and so is this result, which prints
  # the Margin row as a bare `NA` whatever type `year` now holds. A tibble's
  # type header is what shows `year` is still <int>, which is the whole claim
  # being made here.
  dplyr::as_tibble()
```

`.margin_label = NULL` bypasses collision checks. `NA_character_` requests
the same typed missing output but still checks for source missing values and
factor NA levels when `.check_margin_label = TRUE`.

A factor NA level and a missing factor code are different even though both
can print as `<NA>`. Here the second value uses the NA level and the third is
actually missing:

```{r}
factor_status <- structure(
  c(1L, 2L, NA_integer_),
  levels = c("kept", NA_character_),
  class = "factor"
)
is.na(factor_status)
levels(factor_status)

factor_margins <- data.frame(
  status = factor_status,
  value = 1:3
) |>
  summarize_with_margins(
    n = dplyr::n(),
    status_is_margin = grouping_bit(status),
    .grouping = rollup(status),
    .margin_label = NULL
  )

levels(factor_margins$status)
is.na(factor_margins$status)
factor_margins
```

`NULL` preserves the NA level and uses a missing factor code for the Margin
row. Retain `grouping_bit()` or `grouping_id()` because source missing values
and generated Margin values are otherwise indistinguishable. With
`NA_character_`, an existing factor NA level is a structural conflict even
when `.check_margin_label = FALSE`.

For local data, `.check_margin_label = TRUE` prevents an existing value such
as `"Total"` from becoming visually ambiguous. Lazy inputs default to
`.check_margin_label = FALSE` because checking would execute an additional
query. Result row order is unspecified for both local and lazy inputs unless
`.sort` asks for a Margin order; use `arrange()` for any other presentation
order. [Put each subtotal with the rows it
summarizes](https://sayuks.github.io/marginplyr/vignettes/recipes.html)
shows `.sort` at work and what it does not promise. A label is only a
display value; retain `grouping_bit()` or `grouping_id()` whenever source
values can equal the label.

Naming rules, ordered-factor preservation, level positioning, and the complete
eight-case `NA_character_`/`NULL` factor contract live in *Display labels and
grouping identity* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html).

## Advanced dplyr behavior

### Persistent groups as fixed keys

Existing `dplyr::group_by()` columns provide fixed keys when `.by` is not
supplied:

```{r}
grouped_monthly_report <- retail_sales |>
  group_by(year, month) |>
  summarize_with_margins(
    units = sum(units),
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  )

grouped_monthly_report |>
  filter(year == 2026L, month == "Jan")

group_vars(grouped_monthly_report)
```

The calculation is equivalent to the earlier `.by = c(year, month)` call.
Do not combine grouped input with `.by`. A fixed grouping column also cannot
appear in `.grouping`; call `ungroup()` first if that column should be rolled
up instead. Row-wise inputs must likewise be ungrouped explicitly. Local
groups created with `.drop = FALSE` are rejected because empty factor groups
do not have a consistent SQL or lazy-backend equivalent.

Input groups affect the calculation but, except for the explicitly row-wise
nesting verb, do not persist on the result. `summarize_with_margins()`,
`expand_with_margins()`, and `nest_with_margins()` all return ungrouped
output, because a result spanning several grains has no single grouping
hierarchy to keep. `nest_by_with_margins()` is row-wise by every visible key,
grouping-set dimensions included.

This invariant is shared by local data frames and supported lazy backends.
The nesting functions support local data frames and dtplyr because SQL and
Arrow tables cannot contain their list-column result.

*Grouped and row-wise inputs* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
states the rule for all four verbs.

### Intentional differences from `dplyr::summarize()`

Summary expressions use dplyr data masking, but marginplyr narrows the parts
of the official API where arbitrary grouping sets have no portable
interpretation. A summary may not overwrite a fixed key or grouping dimension,
even where the local dplyr backend would allow it, because that would destroy
grouping-set identity. `across()` and `pick()` exclude every column in the
complete grouping plan, not only the ones active in one branch. The
`cur_group*()` family is rejected outright: a branch-local identifier, row
position, or column set has no global meaning once several grouping sets are
combined, so use `grouping_bit()` or `grouping_id()` to identify levels
instead. The refusal is a Package condition naming the replacement, not a
silently different answer:

```{r}
#| label: cur-group-id-rejected
#| must_error: marginplyr_error

retail_sales |>
  summarize_with_margins(
    branch = cur_group_id(),
    .grouping = cube(region, channel)
  )
```

Two familiar conveniences also behave differently. Row order is unspecified
for local and lazy results alike unless `.sort` asks for a Margin order, which
puts each margin row with the rows it summarizes; call `arrange()` for any
other presentation order. And the Margin verbs are not S3 generics: they share
one validated plan, and a private adapter chooses the backend. As in dplyr,
`summarise_with_margins()` and `summarize_with_margins()` are synonyms.

*Relationship to dplyr summaries* in the [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
states each of these rules in full.

For example, `everything()` below selects measures but not `year`, `region`,
or `store`:

```{r}
retail_sales |>
  summarize_with_margins(
    across(
      everything(),
      n_distinct,
      .names = "distinct_{.col}"
    ),
    selected_measures = paste(
      names(pick(everything())),
      collapse = ", "
    ),
    .by = year,
    .grouping = rollup(region, store)
  ) |>
  arrange(year, region, store) |>
  head(n = 6)
```

A summary may return a one-row data frame as in dplyr, but none of its output
columns may be named `year`, `region`, or `store` in this example. Use a new
summary name or rename the input key first.

## Compare a summary with its rollup parent

`share_of_parent()` divides one preceding named numeric scalar summary by its
corresponding value in the immediately less detailed rollup level:

```{r}
parent_report <- retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    revenue_share = share_of_parent(revenue),
    .grouping = rollup(region, store)
  )

parent_report |>
  arrange(region, store)
```

A row of the Grand total set has a Parent share of one. Missing numerators or
denominators and zero denominators produce a missing double value. The helper
requires a rollup and uses internal Grouping identity, so displayed Margin
labels and source missing keys do not determine the parent. For multiple
measures, complete source rules, and the supported `across()` form, see the
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html).

For several measures, use two ordered `across()` expressions: the first
creates ordinary summaries and the second selects those summaries for Parent
shares.

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    across(c(units, revenue), sum),
    across(
      c(units, revenue),
      share_of_parent,
      .names = "{.col}_share"
    ),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)
```

Scale, round, format, or otherwise derive from a Parent share after the Margin
summary:

```{r}
parent_report |>
  mutate(revenue_percent = 100 * revenue_share)
```

## Compare a summary with the whole

`share_of_total()` divides by a row of the Grand total set instead of the
parent level. It is written the same way, and both helpers can appear in one
summary:

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    revenue_share = share_of_parent(revenue),
    revenue_of_total = share_of_total(revenue),
    .grouping = rollup(region, store)
  ) |>
  arrange(region, store)
```

A parent is ambiguous outside a rollup, but a grand total is not: a plan
either contains one or it does not. So `share_of_total()` accepts any
Grouping specification that produces one, including the `cube()` and
`grouping_sets()` forms `share_of_parent()` rejects:

```{r}
retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    revenue_of_total = share_of_total(revenue),
    .grouping = cube(region, store)
  ) |>
  arrange(region, store)
```

A `grouping_sets()` specification qualifies only when it includes an empty
`grouping_set()`; one without it is rejected naming that fix. Everything else
is shared with `share_of_parent()` — eligible sources, the `across()` form,
and the value rules — and the
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
documents both helpers.

Which rows the denominator covers is decided by the fixed `.by` keys. [Compare
each row with the right
total](https://sayuks.github.io/marginplyr/vignettes/recipes.html) works
through that choice.

## Complete absent keys before summarizing

Grouping specifications choose reporting grains; they do not create
unobserved dimension values. marginplyr therefore has no `.complete`, `.fill`,
`.empty`, or `complete_with_margins()` interface. Complete input facts with
ordinary tidyr, dplyr, or dbplyr operations when synthetic facts should
participate in totals, or replace values after the summary when the change is
display-only.

The [Complete absent keys before margins
guide](https://sayuks.github.io/marginplyr/vignettes/completing_keys.html)
shows local `tidyr::complete()` with `nesting()`, explicit key scaffolds,
fact markers for counts and means, same-connection dimension tables,
`dbplyr::copy_inline()`, and explicit `dplyr::copy_to()`.

## Use the same plan with remote data

The grouping plan can be applied to a lazy table without first collecting its
rows. This PostgreSQL simulation shows the native SQL without requiring a
server:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

The result remains lazy. Continue with [Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
for `collect()`, live DuckDB execution, portable `UNION ALL` SQL, backend
verification levels, and practical guidance for inspecting a query.

## Keep the detail rows

### Expand rows with `expand_with_margins()`

`expand_with_margins()` emits one copy of each source row for every grouping
set. It is useful when downstream work needs the rows rather than an
immediate summary. The public name describes the result; SQL `UNION ALL` is
only the portable implementation strategy and this function is unrelated to
`dplyr::union_all()`:

```{r}
january_sales <- retail_sales |>
  filter(year == 2026L, month == "Jan")

expanded <- january_sales |>
  expand_with_margins(
    .grouping = rollup(region, store)
  )

expanded |>
  count(region, store, name = "records")
```

The same verb stays lazy for database tables. Its SQLite SQL makes the
row-expansion branches explicit:

```{r}
#| eval: !expr has_sqlite_simulator

dbplyr::tbl_lazy(
  january_sales,
  con = dbplyr::simulate_sqlite()
) |>
  expand_with_margins(
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

Expanded rows are also the way to compute something one aggregation pass
cannot express, which is a real limit on a database rather than a local one.
[Compute what the summary pass cannot
express](https://sayuks.github.io/marginplyr/vignettes/recipes.html) works
through that case.

### Nest locally or with dtplyr

`nest_with_margins()` keeps one detail table for each visible group:

```{r}
nested_sections <- january_sales |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

# `january_sales` is a plain data frame and the outer class is preserved, so
# `nested_sections` prints its list column as flattened values. A tibble
# prints each nested table as its dimensions instead.
nested_sections |>
  dplyr::as_tibble() |>
  head()

# Inspect the source rows behind the first visible group.
nested_sections$data[[1]]
```

Use `.keep = TRUE` when the grouping keys are also needed inside each detail
table. The nested copies preserve their original values even when the
corresponding outer key is a margin label:

```{r}
nested_with_keys <- january_sales |>
  nest_with_margins(
    .grouping = rollup(region, store),
    .keep = TRUE
  )

company_total <- nested_with_keys |>
  filter(region == "Total", store == "Total")

company_total$data[[1]] |>
  distinct(region, store)
```

A dtplyr input performs the expansion and nesting lazily until `collect()`.
`show_query()` displays the generated data.table expression:

```{r}
#| eval: !expr has_dtplyr

nested_dt <- january_sales |>
  dtplyr::lazy_dt() |>
  nest_with_margins(
    .grouping = rollup(region, store)
  )

show_query(nested_dt)
collect(nested_dt)
```

`nest_by_with_margins()` returns a row-wise data frame, making one calculation
per nested report section straightforward:

```{r}
nested_by <- january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  )

nested_by |>
  mutate(
    records = nrow(data),
    units = sum(data$units),
    revenue = sum(data$revenue)
  )

nested_by$data[[1]]
```

The row-wise result is local even when the input is a dtplyr step:

```{r}
#| eval: !expr has_dtplyr

nested_by_dt <- january_sales |>
  dtplyr::lazy_dt() |>
  nest_by_with_margins(
    .grouping = rollup(region, store)
  )

nested_by_dt |>
  mutate(
    records = nrow(data),
    revenue = sum(data$revenue)
  )

nested_by_dt$data[[1]]
```

Because `nest_by_with_margins()` collects before making the result row-wise,
its return value no longer has a lazy query to display. Use the
`nest_with_margins()` form immediately above with `show_query()` to inspect the
same dtplyr expansion and nesting expression before collection.

List columns are why the two nesting verbs support local data frames and
dtplyr steps rather than SQL database tables.

#### Relationship to tidyr and dplyr

The margin-aware verbs deliberately share the main result shapes of
`tidyr::nest()` and `dplyr::nest_by()`, but they are not drop-in replacements.
Both select their keys with `.by` plus `.grouping` rather than data-masking
`...`, and both take a logical `.keep`. `nest_with_margins()` always nests
every non-key column and does not implement `...`, multiple list columns, or
`.names_sep`. `nest_by_with_margins()` collects a dtplyr result before making
it row-wise, because a row-wise data frame is a local object.

Sharing a result shape is not upstream compatibility; the differences are part
of the margin-aware API. *Relationship to tidyr and dplyr* in the
[`nest_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/nest_with_margins.html)
states both contracts in full.

For grouped input, existing grouping columns become fixed keys as though they
had been supplied through `.by`. Unlike `tidyr::nest()`, `nest_with_margins()`
does not preserve that grouping metadata on its output. Row order is
unspecified; call `arrange()` when the outer keys need a presentation order.

Both functions support `.keep`. With `.keep = TRUE`, nested tables retain the
original pre-margin key values. For example, the outer key of an
`East / Total` subtotal displays `"Total"`, while the rows inside contain
their original store names. For `nest_by_with_margins()`, this preserves the
meaning of dplyr's “keep the grouping columns” rule after a dimension has been
rolled up. For `nest_with_margins()`, it corresponds to explicitly selecting
grouping keys with `tidyr::nest(data = everything(), .by = ...)`. marginplyr
uses a logical argument instead of the full `...` interface because its
grouping dimensions must remain visible outside to distinguish detail,
subtotal, and total rows.

Nesting accepts `.duplicates = "error"` or `"drop"`, but rejects `"keep"`.
Unlike a summary or row expansion, two duplicate nested groups have no
visible set identifier and would therefore be indistinguishable. Internal set
identifiers and `.keep` copies use collision-free temporary names, so no input
column name is reserved.

The list column is a regular list of data frames; a
`vctrs::list_of` subclass is not promised. On an empty ungrouped input,
`nest_with_margins()` follows `tidyr::nest()` and returns zero outer rows.
`nest_by_with_margins()` follows `dplyr::nest_by()` and returns one row
containing the empty data frame when there are no keys.

`dplyr::nest_by()` is currently not a stable upstream API and may eventually
be deprecated in favor of `tidyr::nest(.by = )`. marginplyr nevertheless
keeps its row-wise wrapper because per-margin modeling and reporting are
useful, and implements it through the same nesting core rather than a separate
semantic path.

## Optional Quarto tabsets with quartabs

[`quartabs`](https://sayuks.github.io/quartabs/) can turn the nested report
sections into Quarto tabs. This compact example creates one tab per region
plus a company-total tab.

When this guide is opened as an installed package vignette, Quarto's
lightweight vignette format does not provide interactive tabsets. The same
content is displayed sequentially instead. View the
[package-site version of this guide](https://sayuks.github.io/marginplyr/vignettes/get_started.html#optional-quarto-tabsets-with-quartabs)
to use the interactive tabs.

The chunk option `results: asis` is required because `render_tabset()` emits
Quarto tabset markup.

```{r}
#| results: asis
#| eval: !expr has_tabsets

tabbed_report <- january_sales |>
  nest_by_with_margins(
    .grouping = rollup(region)
  ) |>
  mutate(
    # Order the company total before regional tabs in the rendered report.
    region = factor(region, levels = c("Total", "East", "West")),
    report = list(
      data |>
        summarize(
          records = n(),
          units = sum(units),
          revenue = sum(revenue)
        ) |>
        knitr::kable()
    )
  ) |>
  ungroup()

tabbed_report |>
  quartabs::render_tabset(
    tabset_vars = region,
    output_vars = report
  )
```

The `"Total"` tab is created by the grouping plan. The ordered factor makes it
the first tab; `quartabs` only handles presentation, while marginplyr defines
which source rows belong to that total.
