---
title: Grouping identity
vignette: >
  %\VignetteIndexEntry{Grouping identity}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

A Margin result combines rows from several Grouping sets. Visible values such
as `"Total"` are useful for display, but they do not identify which Grouping
set produced a row. marginplyr provides Grouping set identifiers and
structural Grouping identifiers for that purpose.

Read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html) first
if Grouping specifications are new to you. This article focuses on the
difference between plan position, absence patterns, and displayed values.

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

## Four related values

| Interface | Meaning | Values | Duplicate Grouping sets |
|---|---|---|---|
| `.id = "name"` | One-based position of a Grouping-set occurrence in the resolved Grouping plan | `1L`, `2L`, ... | Distinct when `.duplicates = "keep"` |
| `inspect_grouping()$set_id` | The same occurrence position, inspected before a Margin operation | `1L`, `2L`, ... | Distinct when `.duplicates = "keep"` |
| `grouping_bit(x)` | Whether one chosen grouping dimension is absent | `0L` when included, `1L` when absent | The same for identical absence patterns |
| `grouping_id(...)` | Bit mask formed from the chosen Grouping bits | A non-negative integer through 31 chosen dimensions | The same for identical absence patterns |

The first two values describe *where an occurrence appears in one ordered
Grouping plan*. The last two describe *which dimensions are absent from a
row*. A Grouping set identifier can therefore distinguish repeated identical
sets; a Grouping identifier deliberately cannot.

The correspondence between `.id` and `set_id` holds for the same resolved
`.by`, `.grouping`, and `.duplicates` arguments. Reordering or deduplicating a
Grouping specification can change both Grouping set identifiers.

## Why a rollup skips Grouping identifier 2

Inspect a two-dimension rollup:

```{r}
rollup_plan <- inspect_grouping(
  .data = retail_sales,
  .grouping = rollup(region, store)
)

rollup_plan
```

The last argument to `grouping_id(region, store)` is the least-significant
bit:

| Grouping set | `region` bit | `store` bit | Binary mask | Grouping identifier |
|---|---:|---:|---:|---:|
| `(region, store)` | 0 | 0 | `00` | 0 |
| `(region)` | 0 | 1 | `01` | 1 |
| `()` | 1 | 1 | `11` | 3 |

Identifier `2` would be binary `10`: `region` absent while `store` remains.
That is not part of the hierarchy declared by `rollup(region, store)`, so the
identifier is absent. Grouping identifiers encode absence; they are not
required to be consecutive.

The Grouping set identifiers are a different sequence:

```{r}
rollup_result <- retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    region_bit = grouping_bit(region),
    store_bit = grouping_bit(store),
    absence = grouping_id(region, store),
    .grouping = rollup(region, store),
    .id = "occurrence"
  )

rollup_result |>
  distinct(occurrence, region_bit, store_bit, absence) |>
  arrange(occurrence)
```

Here `.id` is `1L`, `2L`, and `3L`, while `absence` is `0L`, `1L`, and `3L`.

## A cube includes every mask

`cube(region, store)` contains every subset, including `(store)`:

```{r}
cube_plan <- inspect_grouping(
  .data = retail_sales,
  .grouping = cube(region, store)
)

cube_plan
```

Its absence masks are `00`, `01`, `10`, and `11`, so its Grouping identifiers
are `0`, `1`, `2`, and `3`.

## Occurrence does not promise row order

`inspect_grouping()` returns rows in Grouping-plan order, but Margin-operation
results have no implicit physical order. `.id` records which occurrence
produced a row; it does not request an `ORDER BY` or sort rows within an
occurrence.

Use `arrange()` explicitly whenever presentation order matters:

```{r}
rollup_result |>
  arrange(occurrence, region, store)
```

Do not use `.id` as a durable business key between results made from
differently ordered or deduplicated Grouping specifications.

## Inspect a Grouping plan as data

`inspect_grouping()` returns an ordinary local, ungrouped tibble. It does not
aggregate source rows, even when `.data` is a lazy table. This makes the plan
easy to view in an IDE, transform with dplyr, export, or snapshot-test:

```{r}
class(rollup_plan)
group_vars(rollup_plan)

rollup_plan |>
  filter(grouping_id != 0L) |>
  select(set_id, included, omitted, grouping_id)
```

The default `.format = "text"` avoids opaque list-column cells in IDE table
viewers. Use `.format = "list"` when exact names or programmable vectors
matter:

```{r}
rollup_list <- inspect_grouping(
  .data = retail_sales,
  .grouping = rollup(region, store),
  .format = "list"
)

rollup_list$omitted
rollup_list$grouping_bits
```

The list format stores `fixed`, `included`, and `omitted` as character
list-columns and `grouping_bits` as a named integer-vector list-column. The
text form is for display, not serialization.

`grouping_id` uses an R integer and is available through 31 variable
dimensions. Beyond that limit it is `NA_integer_`; `grouping_bits` remains
complete:

```{r}
wide_data <- as.data.frame(
  stats::setNames(rep(list(1), 32L), sprintf("x%02d", 1:32))
)
wide_plan <- inspect_grouping(
  .data = wide_data,
  .grouping = rollup(everything()),
  .format = "list"
)

wide_plan |>
  slice(1L) |>
  transmute(
    number_of_bits = lengths(grouping_bits),
    grouping_id
  )
```

This is a backend-independent Grouping plan, not a database execution plan.
Use `show_query()` for generated SQL and database-native tooling for optimizer
plans:

```{r}
dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
) |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = rollup(region, store)
  ) |>
  show_query()
```

## Margin labels are display values

A Margin label replaces an omitted dimension for display. Changing the label
does not change `.id`, Grouping bits, or a Grouping identifier:

```{r}
labeled_result <- retail_sales |>
  filter(year == 2026L, month == "Jan") |>
  summarize_with_margins(
    revenue = sum(revenue),
    absence = grouping_id(region, store),
    .grouping = rollup(region, store),
    .margin_label = c(
      region = "All regions",
      store = "All stores"
    ),
    .id = "occurrence"
  )

labeled_result |>
  arrange(occurrence, region, store)
```

If a source value can equal a display label, retain `.id`, `grouping_bit()`,
or `grouping_id()` instead of guessing from the visible value.

## Three kinds of apparent missingness

Factors make three distinct situations visible:

1. An **NA factor level** belongs to the factor domain and can be used by an
   observation whose factor code is not missing.
2. A **source missing value** has a missing factor code.
3. A **typed-missing Margin value** marks a dimension omitted from a Grouping
   set when `.margin_label = NULL` or `NA_character_`.

The first two can both print as `<NA>`:

```{r}
status <- structure(
  c(1L, 2L, NA_integer_),
  levels = c("kept", NA_character_),
  class = "factor"
)

levels(status)
is.na(status)
```

`NULL` preserves the NA level and uses a missing factor code for the Margin
row:

```{r}
factor_result <- data.frame(status, value = 1:3) |>
  summarize_with_margins(
    records = dplyr::n(),
    status_is_margin = grouping_bit(status),
    .grouping = rollup(status),
    .margin_label = NULL
  )

levels(factor_result$status)
is.na(factor_result$status)
factor_result |>
  arrange(status_is_margin, status)
```

The structural flag, rather than the printed value, distinguishes a source
missing value from the typed-missing Margin value. See the
[`summarize_with_margins()` reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html#display-labels-and-grouping-identity)
for the complete eight-case `NA_character_`/`NULL` factor table, ordered-factor
preservation, unused-level checks, and label positioning.

## Parent lookup is structural

`share_of_parent()` matches a row to its immediate less detailed rollup parent
with internal Grouping identity and missing-safe key identity. It does not join
on displayed Margin labels, and it does not use `.id` as a parent key:

```{r}
parent_result <- data.frame(
  region = c("East", "East"),
  store = c("A", NA_character_),
  revenue = c(80, 20)
) |>
  summarize_with_margins(
    revenue = sum(revenue),
    revenue_share = share_of_parent(revenue),
    absence = grouping_id(region, store),
    .grouping = rollup(region, store),
    .margin_label = NULL
  )

parent_result |>
  arrange(absence, region, store)
```

The included source `NA` store and the subtotal's omitted `store` may display
the same value, but their Grouping identities differ and Parent lookup remains
unambiguous. See the
[`share_of_parent()` reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
for its value, source-summary, ordering, and `across()` rules.
