---
title: Recipes for common reporting tasks
vignette: >
  %\VignetteIndexEntry{Recipes for common reporting tasks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

The other guides are organized by concept. This one is organized by task: each
section states a reporting question, answers it with one worked pipeline, and
stops. Read them in any order and skip the rest — no recipe uses a value
another recipe defined, so every one starts from `retail_sales` or from its
own lazy table.

Read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html) first
if Grouping specifications are new to you. Each recipe links to the concept
guide that explains why its answer works.

## Installation

Install marginplyr from GitHub:

```{r}
#| eval: false

install.packages("pak")
pak::pkg_install("sayuks/marginplyr")

# Used for the sections of this guide that execute against a database.
install.packages(c("DBI", "duckdb"))
```

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. Three recipes execute against DuckDB, so
their chunks evaluate only when DuckDB is installed at the version marginplyr
asks for. Where a package is missing or too old the code is still shown, but
its output is absent. Everything else runs either locally or against
`dbplyr::simulate_postgres()`, which needs no database and no optional package.

```{r}
#| include: false

# `marginplyr_suggest_available()` answers what a bare installation check
# could not: whether the installed package satisfies the version DESCRIPTION
# requires. AGENTS.md is authoritative for which guards must go through it.
source(system.file("suggests", "guard.R", package = "marginplyr"))

has_duckdb <- marginplyr_suggest_available("DBI") &&
  marginplyr_suggest_available("duckdb")

# Chunks marked with `must_error` show a call that is meant to fail, so the
# reader sees the real diagnostic rather than a quotation of one. AGENTS.md is
# authoritative for what the option does and why it exists.
source(system.file("vignette-hooks", "must-error.R", package = "marginplyr"))
```

## Put each subtotal with the rows it summarizes

By default a Margin result promises no particular arrangement, so subtotals
can land anywhere:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store)
  )
```

`.sort = "last"` asks for a Margin order instead. Within each fixed `.by` key
every dimension contributes its Grouping bit before its own value, so each
subtotal sits directly beneath the rows it summarizes:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .sort = "last"
  )
```

`"first"` puts each subtotal above its rows instead. Neither choice looks at
`.margin_label`, so the placement does not change if the label does, and it
does not depend on the locale. The `NA` store sorts after the named stores but
still ahead of the region subtotal: each dimension contributes its Grouping
bit before its missingness, so a margin outranks a missing value rather than
competing with one.

**A Margin order belongs to the object the verb returned, not to anything
derived from it.** Any later verb is free to reorder, and on a lazy backend
one usually does — see [Name the grouping set on every
row](#name-the-grouping-set-on-every-row) for a join that drops it silently.
When some other presentation order is what you want, write the key yourself:
[Write the sort key yourself](#write-the-sort-key-yourself) shows it as an
`arrange()` you control.

[Grouping
identity](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html)
explains why Grouping-plan order and a Margin order answer different
questions.

## Write the sort key yourself

A Margin order leaves every dimension's values ascending, so a report that
wants the newest year at the top asks for the same key with one term changed.
`grouping_bit()` puts the Grouping bits the key reads into the result, and the
`arrange()` is then the key above with `desc()` on the term you are reversing:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    yb = grouping_bit(year),
    rb = grouping_bit(region),
    .grouping = rollup(year, region)
  ) |>
  arrange(yb, is.na(year), desc(year), rb, is.na(region), region) |>
  select(-yb, -rb)
```

2026 sits above 2025 and each subtotal is still under its own rows, because
everything ahead of the reversed term is what it was.

An order no sort key can express is written the same way, with one more
column. Ordering regions by revenue, and the stores inside each region by
revenue, needs each region's own revenue on every one of its rows; that number
is already in the result as the region's subtotal row, so a grouped `mutate()`
can put it there, and the `sb` term keeps each subtotal under its own stores:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    rb = grouping_bit(region),
    sb = grouping_bit(store),
    .grouping = rollup(region, store)
  ) |>
  mutate(
    .by = region,
    region_revenue = max(ifelse(sb == 1, revenue, NA), na.rm = TRUE)
  ) |>
  arrange(rb, desc(region_revenue), region, sb, desc(revenue), store) |>
  select(region, store, revenue)
```

**On a lazy backend, dropping a helper column can cost the order.** Whether
it does is decided by the query dbplyr builds, which this package does not own
— but you are told when it happens, either by dbplyr warning that it discarded
an `ORDER BY` or by the database refusing a query that orders on a column no
longer in it.

A Grouping bit usually survives, because the bits of a `rollup()` are computed
in the query underneath the one the ordering sits in. The reversed-year recipe runs on
a database unchanged:

```{r}
#| eval: !expr has_duckdb

# `shared_home = FALSE` keeps DuckDB's extension cache and stored secrets in
# the session's temporary directory rather than in `~/.duckdb`.
con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
duckdb_sales <- dplyr::copy_to(con, retail_sales, "retail_sales")

duckdb_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    yb = grouping_bit(year),
    rb = grouping_bit(region),
    .grouping = rollup(year, region)
  ) |>
  arrange(yb, is.na(year), desc(year), rb, is.na(region), region) |>
  select(-yb, -rb) |>
  collect()
```

`region_revenue` does not survive it. You computed it after the summary, so it
is in the same query as the ordering rather than underneath it, and dropping it
leaves an `ORDER BY` naming a column that is gone:

```{r}
#| label: dropping-a-computed-sort-term
#| must_error: true
#| eval: !expr has_duckdb

duckdb_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    rb = grouping_bit(region),
    sb = grouping_bit(store),
    .grouping = rollup(region, store)
  ) |>
  mutate(
    .by = region,
    region_revenue = max(ifelse(sb == 1, revenue, NA), na.rm = TRUE)
  ) |>
  arrange(rb, desc(region_revenue), region, sb, desc(revenue), store) |>
  select(region, store, revenue) |>
  collect()
```

Keep the column and the result stays lazy and stays ordered, at the cost of
carrying it:

```{r}
#| eval: !expr has_duckdb

duckdb_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    rb = grouping_bit(region),
    sb = grouping_bit(store),
    .grouping = rollup(region, store)
  ) |>
  mutate(
    .by = region,
    region_revenue = max(ifelse(sb == 1, revenue, NA), na.rm = TRUE)
  ) |>
  arrange(rb, desc(region_revenue), region, sb, desc(revenue), store) |>
  collect()
```

`collect()` before the `select()` is the other way, and gives up laziness to
get a report with no helper columns in it.

```{r}
#| include: false
#| eval: !expr has_duckdb

DBI::dbDisconnect(con, shutdown = TRUE)
```

The [`summarize_with_margins()`
reference](https://sayuks.github.io/marginplyr/man/summarize_with_margins.html)
states the whole key under *Margin order*.

## Compare each row with the right total

`share_of_total()` divides a summary by the same summary on the Grand total
set. Which rows that set covers is decided by the fixed `.by` keys, and that
is the whole of the choice. With `.by = year`, each year is its own
denominator:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    share_of_year = share_of_total(revenue),
    .by = year,
    .grouping = rollup(region),
    .sort = "last"
  )
```

The region rows inside each year add up to one, and 2025 contributes nothing
to 2026's denominator. The year's own row is the denominator, so its share is
one by definition rather than by addition. Move the dimension out of `.by` and
into the Grouping specification, and the same helper measures against the
whole table instead:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    share_overall = share_of_total(revenue),
    .grouping = rollup(year, region),
    .sort = "last"
  )
```

Now every row is a share of all four region-years together, and the year rows
are subtotals rather than partitions. Both calls are correct; they answer
different questions, and `.by` is what distinguishes them.

For a denominator one rollup level up rather than at the top, use
`share_of_parent()`. The
[`share_of_parent()`
reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
documents both helpers, and [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
compares them side by side.

## Keep only one grouping set

Filtering on a Margin label is unsafe: a source value can equal the label, and
the label can be changed or set to `NULL`. Filter on Grouping identity
instead. The obstacle is knowing which value to ask for — and guessing the bit
order is easy to get backwards.

Do not guess. `inspect_grouping()` prints the answer for every set in the plan
before any data is summarized:

```{r}
inspect_grouping(
  .data = retail_sales,
  .grouping = cube(region, channel)
)
```

Read `grouping_bits` beside `grouping_id`: the last dimension is the least
significant bit, so `channel` alone being omitted is `1`, `region` alone is
`2`, and both is `3`. Retain the identifier in the summary and filter on it:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    absence = grouping_id(region, channel),
    .grouping = cube(region, channel),
    .sort = "last"
  ) |>
  filter(absence == 1L)
```

Two steps are required because `grouping_id()` and `grouping_bit()` are
contextual summary helpers. They read the Grouping set that produced a row,
which only the summary knows, so calling one inside `filter()` is rejected:

```{r}
#| label: grouping-bit-in-filter
#| must_error: marginplyr_error

retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(region, channel)
  ) |>
  filter(grouping_bit(channel) == 1L)
```

For a single dimension, `grouping_bit()` avoids the bit arithmetic altogether:
retain one flag column per dimension and filter on the flags. [Grouping
identity](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html)
compares all four identity values and states their duplicate-occurrence rules.

## Name the grouping set on every row

`.id` numbers the Grouping-set occurrence behind each row, and
`inspect_grouping()` returns one row per occurrence keyed by the same number.
Joining them attaches whatever name you care to write.

Build the name on the plan first. `inspect_grouping()` returns an ordinary
tibble with no custom class ([ADR 0013][adr13]), so ordinary dplyr applies,
and cutting
it down to two columns keeps the plan's five description columns out of the
report:

```{r}
sales_levels <- retail_sales |>
  inspect_grouping(
    .by = year,
    .grouping = rollup(region, store)
  ) |>
  mutate(
    level = case_when(
      omitted == "()" ~ "Store detail",
      omitted == "(store)" ~ "Region subtotal",
      .default = "All regions"
    )
  ) |>
  select(set_id, level)

sales_levels
```

Then join it onto the result:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  ) |>
  left_join(sales_levels, by = c(set = "set_id"))
```

### The same join against a lazy table

`inspect_grouping()` is local even when `.data` is not: it reads captured
column metadata and the Grouping plan, never the source rows
([ADR 0013][adr13]). A lazy result and a local tibble have no common source,
so dbplyr refuses the join outright:

[adr13]: https://github.com/sayuks/marginplyr/blob/main/design/adr/0013-inspect-grouping-plans-as-ordinary-tibbles.md

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_report <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  )
```

```{r}
#| label: join-without-copy
#| must_error: true

postgres_report |>
  left_join(sales_levels, by = c(set = "set_id"))
```

`copy = "inline"` compiles the plan into the query as a `VALUES` list, which
suits a table holding one row per grouping set and needs no permission to
write one:

```{r}
postgres_report |>
  left_join(sales_levels, by = c(set = "set_id"), copy = "inline")
```

Two conditions come with that route, and neither announces itself.

**The Margin order does not survive the join.** As [Put each subtotal with the
rows it summarizes](#put-each-subtotal-with-the-rows-it-summarizes) put it, a
Margin order is a property of the object the verb returned, not of anything
derived from it. A simulated connection renders SQL without running it, so
seeing this needs a live database:

```{r}
#| eval: !expr has_duckdb

con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
duckdb_sales <- dplyr::copy_to(con, retail_sales, "retail_sales")

duckdb_report <- duckdb_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  )

duckdb_report |>
  left_join(sales_levels, by = c(set = "set_id"), copy = "inline") |>
  collect()
```

The subtotals are scattered and the years interleave. A database is entitled
to ignore an `ORDER BY` buried in a subquery, and this one does; the only
warning is dbplyr's generic *ORDER BY is ignored in subqueries without LIMIT*,
which names nothing in marginplyr, while `show_query()` goes on printing an
`# Ordered by:` line for an order that is no longer honored. `.sort` is an
argument of the Margin verb, so it cannot be applied again afterwards — order
with `arrange()` after the join, or collect first as below.

**The plan has to be in the default `.format = "text"`.** The list columns of
`.format = "list"` have no SQL type, and the failure surfaces while dbplyr
renders the query, without naming `.format`:

```{r}
list_levels <- inspect_grouping(
  .data = postgres_sales,
  .by = year,
  .grouping = rollup(region, store),
  .format = "list"
)
```

```{r}
#| label: join-list-format
#| must_error: true

postgres_report |>
  left_join(list_levels, by = c(set = "set_id"), copy = "inline") |>
  show_query()
```

Prefer `collect()` before the join whenever the result is a report you are
about to read. It is the same shape as the local join at the top of this
recipe: the plan is one row per grouping set and a finished summary is already
small, so nothing is gained by making the database perform the join — and
collecting first keeps the Margin order, allows `.format = "list"`, and needs
no `copy` argument at all:

```{r}
#| eval: !expr has_duckdb

duckdb_report |>
  collect() |>
  left_join(sales_levels, by = c(set = "set_id"))
```

Same query, same plan, same join — and every subtotal is back under the rows
it summarizes.

```{r}
#| include: false
#| eval: !expr has_duckdb

DBI::dbDisconnect(con, shutdown = TRUE)
```

[Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
covers query shape and execution for lazy inputs generally.

## Compute what the summary pass cannot express

A summary expression can do more than call one aggregate. On a local data
frame dplyr hands it the whole group, so a calculation that needs both a
group-level value and the rows behind it works directly — here, the share of
revenue coming from rows above their own set's mean:

```{r}
retail_sales |>
  summarize_with_margins(
    above_mean_share =
      sum(revenue[revenue > mean(revenue)]) / sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .sort = "last"
  )
```

Against a database the same expression has to compile to SQL aggregates, and
this one does not: the comparison needs an average that is itself an aggregate
over the same group, which no dialect allows inside another aggregate. The
error comes from the database, with its own wording:

```{r}
#| eval: !expr has_duckdb

con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
duckdb_sales <- dplyr::copy_to(con, retail_sales, "retail_sales")
```

```{r}
#| label: nested-aggregate
#| must_error: true
#| eval: !expr has_duckdb

duckdb_sales |>
  summarize_with_margins(
    above_mean_share =
      sum(revenue[revenue > mean(revenue)]) / sum(revenue),
    .by = year,
    .grouping = rollup(region, store)
  ) |>
  collect()
```

`expand_with_margins()` is the way through. It emits one copy of each source
row for every grouping set, which turns the one summary pass into ordinary
rows you can take as many passes over as the calculation needs — a window
function for the group's mean, then an aggregate over the comparison. It stays
lazy throughout:

```{r}
#| eval: !expr has_duckdb

duckdb_sales |>
  expand_with_margins(
    .by = year,
    .grouping = rollup(region, store),
    .id = "set"
  ) |>
  group_by(year, region, store, set) |>
  mutate(set_mean = mean(revenue, na.rm = TRUE)) |>
  summarize(
    above_mean_share =
      sum(ifelse(revenue > set_mean, revenue, 0), na.rm = TRUE) /
      sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  collect()
```

```{r}
#| include: false
#| eval: !expr has_duckdb

DBI::dbDisconnect(con, shutdown = TRUE)
```

Every value matches the local calculation above. The two tables are not laid
out alike — this one carries the `set` column and is in no Margin order,
because the pipeline ends in an ordinary `summarize()` rather than in a Margin
verb — so compare them by key rather than row by row. The cost is in the row
count:
expansion produces one copy of the input per grouping set, so a plan with many
sets multiplies the rows the database has to scan. Reach for this when the
aggregation pass genuinely cannot express the calculation, not to avoid
writing a summary.

[Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
shows the SQL `expand_with_margins()` generates on native and fallback
backends.
