---
title: Recipes for common reporting tasks
vignette: >
  %\VignetteIndexEntry{Recipes for common reporting tasks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    code-tools: true
    toc: true
    toc-expand: true
    number-sections: true
date: last-modified
---

The other guides are organized by concept. This one is organized by task: each
section states a reporting question, answers it with one worked pipeline, and
stops. Read them in any order and skip the rest — no recipe uses a value
another recipe defined, so every one starts from `retail_sales` or from its
own lazy table.

Read [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html) first
if Grouping specifications are new to you. Each recipe links to the concept
guide that explains why its answer works.

```{r setup}
#| message: false

library(marginplyr)
library(dplyr)
```

Only dplyr and dbplyr are required. One recipe executes against DuckDB, so its
chunks evaluate only when DuckDB is installed. Where a package is missing the
code is still shown, but its output is absent. Everything else runs either
locally or against `dbplyr::simulate_postgres()`, which needs no database and
no optional package.

```{r}
#| include: false

has_duckdb <- requireNamespace("DBI", quietly = TRUE) &&
  requireNamespace("duckdb", quietly = TRUE)
```

```{r}
#| include: false

# Several recipes below show a call that is meant to fail, so the reader sees
# the real diagnostic rather than a quotation of one. Quarto's `error: true`
# permits an error without requiring one, which would let this prose outlive
# the behavior it describes. `must_error: true` adds the missing half: it
# implies `error: true` and halts the render if the chunk stops failing.
# See AGENTS.md.
local({
  knitr::opts_hooks$set(must_error = function(options) {
    if (isTRUE(options$must_error)) options$error <- TRUE
    options
  })
  default_evaluate <- knitr::knit_hooks$get("evaluate")
  knitr::knit_hooks$set(evaluate = function(...) {
    res <- default_evaluate(...)
    opts <- knitr::opts_current$get()
    if (isTRUE(opts$must_error) &&
          !any(vapply(res, inherits, logical(1), "error"))) {
      stop(
        "Chunk `", opts$label, "` is marked `must_error: true` but ",
        "completed without raising an error.",
        call. = FALSE
      )
    }
    res
  })
})
```

## Put each subtotal with the rows it summarizes

By default a Margin result promises no particular arrangement, so subtotals
can land anywhere:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store)
  )
```

`.sort = "last"` asks for a Margin order instead. Within each fixed `.by` key
every dimension contributes its Grouping bit before its own value, so each
subtotal sits directly beneath the rows it summarizes:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .sort = "last"
  )
```

`"first"` puts each subtotal above its rows instead. Neither choice looks at
`.margin_label`, so the placement does not change if the label does, and it
does not depend on the locale. The `NA` store keeps its own position because
missing values sort last wherever they appear, independently of the margins.

**A Margin order belongs to the object the verb returned, not to anything
derived from it.** Any later verb is free to reorder, and on a lazy backend
one usually does — see [Name the grouping set on every
row](#name-the-grouping-set-on-every-row) for a join that drops it silently.
When some other presentation order is what you want, `arrange()` after the
fact and do not ask for a Margin order at all.

[Grouping
identity](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html)
explains why Grouping-plan order and a Margin order answer different
questions.

## Compare each row with the right total

`share_of_total()` divides a summary by the same summary on the Grand total
set. Which rows that set covers is decided by the fixed `.by` keys, and that
is the whole of the choice. With `.by = year`, each year is its own
denominator:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    share_of_year = share_of_total(revenue),
    .by = year,
    .grouping = rollup(region),
    .sort = "last"
  )
```

The shares inside each year add up to one, and 2025 contributes nothing to
2026's denominator. Move the dimension out of `.by` and into the Grouping
specification, and the same helper measures against the whole table instead:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    share_overall = share_of_total(revenue),
    .grouping = rollup(year, region),
    .sort = "last"
  )
```

Now every row is a share of all four region-years together, and the year rows
are subtotals rather than partitions. Both calls are correct; they answer
different questions, and `.by` is what distinguishes them.

For a denominator one rollup level up rather than at the top, use
`share_of_parent()`. The
[`share_of_parent()`
reference](https://sayuks.github.io/marginplyr/man/share_of_parent.html)
documents both helpers, and [Get
started](https://sayuks.github.io/marginplyr/vignettes/get_started.html)
compares them side by side.

## Keep only one grouping set

Filtering on a Margin label is unsafe: a source value can equal the label, and
the label can be changed or set to `NULL`. Filter on Grouping identity
instead. The obstacle is knowing which value to ask for — and guessing the bit
order is easy to get backwards.

Do not guess. `inspect_grouping()` prints the answer for every set in the plan
before any data is summarized:

```{r}
inspect_grouping(
  .data = retail_sales,
  .grouping = cube(region, channel)
)
```

Read `grouping_bits` beside `grouping_id`: the last dimension is the least
significant bit, so `channel` alone being omitted is `1`, `region` alone is
`2`, and both is `3`. Retain the identifier in the summary and filter on it:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    absence = grouping_id(region, channel),
    .grouping = cube(region, channel),
    .sort = "last"
  ) |>
  filter(absence == 1L)
```

Two steps are required because `grouping_id()` and `grouping_bit()` are
contextual summary helpers. They read the Grouping set that produced a row,
which only the summary knows, so calling one inside `filter()` is rejected:

```{r}
#| label: grouping-bit-in-filter
#| must_error: true

retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .grouping = cube(region, channel)
  ) |>
  filter(grouping_bit(channel) == 1L)
```

For a single dimension, `grouping_bit()` avoids the bit arithmetic altogether:
retain one flag column per dimension and filter on the flags. [Grouping
identity](https://sayuks.github.io/marginplyr/vignettes/grouping_identity.html)
compares all four identity values and states their duplicate-occurrence rules.

## Name the grouping set on every row

`.id` numbers the Grouping-set occurrence behind each row, and
`inspect_grouping()` returns one row per occurrence keyed by the same number.
Joining them attaches whatever name you care to write.

Build the name on the plan first. `inspect_grouping()` returns an ordinary
tibble with no custom class (ADR 0013), so ordinary dplyr applies, and cutting
it down to two columns keeps the plan's five description columns out of the
report:

```{r}
sales_levels <- retail_sales |>
  inspect_grouping(
    .by = year,
    .grouping = rollup(region, store)
  ) |>
  mutate(
    level = case_when(
      omitted == "()" ~ "Store detail",
      omitted == "(store)" ~ "Region subtotal",
      .default = "All regions"
    )
  ) |>
  select(set_id, level)

sales_levels
```

Then join it onto the result:

```{r}
retail_sales |>
  summarize_with_margins(
    revenue = sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  ) |>
  left_join(sales_levels, by = c(set = "set_id"))
```

### The same join against a lazy table

`inspect_grouping()` is local even when `.data` is not: it reads captured
column metadata and the Grouping plan, never the source rows (ADR 0013). A
lazy result and a local tibble have no common source, so dbplyr refuses the
join outright:

```{r}
postgres_sales <- dbplyr::tbl_lazy(
  retail_sales,
  con = dbplyr::simulate_postgres()
)

postgres_report <- postgres_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  )
```

```{r}
#| label: join-without-copy
#| must_error: true

postgres_report |>
  left_join(sales_levels, by = c(set = "set_id"))
```

`copy = "inline"` compiles the plan into the query as a `VALUES` list, which
suits a table holding one row per grouping set and needs no permission to
write one:

```{r}
postgres_report |>
  left_join(sales_levels, by = c(set = "set_id"), copy = "inline")
```

Two conditions come with that route, and neither announces itself.

**The Margin order does not survive the join.** As the previous recipe put it,
a Margin order is a property of the object the verb returned, not of anything
derived from it. Here the database is entitled to ignore an `ORDER BY` buried
in a subquery, and does; the only warning is dbplyr's generic *ORDER BY is
ignored in subqueries without LIMIT*, which names nothing in marginplyr, while
`show_query()` goes on printing an `# Ordered by:` line for an order that is
no longer honored. `.sort` is an argument of the Margin verb, so it cannot be
applied again afterwards — order with `arrange()` after the join, or collect
first as below.

**The plan has to be in the default `.format = "text"`.** The list columns of
`.format = "list"` have no SQL type, and the failure surfaces while dbplyr
renders the query, without naming `.format`:

```{r}
list_levels <- inspect_grouping(
  .data = postgres_sales,
  .by = year,
  .grouping = rollup(region, store),
  .format = "list"
)
```

```{r}
#| label: join-list-format
#| must_error: true

postgres_report |>
  left_join(list_levels, by = c(set = "set_id"), copy = "inline") |>
  show_query()
```

Prefer `collect()` before the join whenever the result is a report you are
about to read. The plan is one row per grouping set and a finished summary is
already small, so nothing is gained by making the database perform the join —
and collecting first keeps the Margin order, allows `.format = "list"`, and
needs no `copy` argument at all:

```{r}
#| eval: !expr has_duckdb

con <- DBI::dbConnect(duckdb::duckdb())
duckdb_sales <- dplyr::copy_to(con, retail_sales, "retail_sales")

duckdb_sales |>
  summarize_with_margins(
    revenue = sum(revenue, na.rm = TRUE),
    .by = year,
    .grouping = rollup(region, store),
    .id = "set",
    .sort = "last"
  ) |>
  collect() |>
  left_join(sales_levels, by = c(set = "set_id"))
```

```{r}
#| include: false
#| eval: !expr has_duckdb

DBI::dbDisconnect(con, shutdown = TRUE)
```

[Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
covers query shape and execution for lazy inputs generally.

## Compute what the summary pass cannot express

A summary expression can do more than call one aggregate. On a local data
frame dplyr hands it the whole group, so a calculation that needs both a
group-level value and the rows behind it works directly — here, the share of
revenue coming from rows above their own set's mean:

```{r}
retail_sales |>
  summarize_with_margins(
    above_mean_share =
      sum(revenue[revenue > mean(revenue)]) / sum(revenue),
    .by = year,
    .grouping = rollup(region, store),
    .sort = "last"
  )
```

Against a database the same expression has to compile to SQL aggregates, and
this one does not: the comparison needs an average that is itself an aggregate
over the same group, which no dialect allows inside another aggregate. The
error comes from the database, with its own wording:

```{r}
#| eval: !expr has_duckdb

con <- DBI::dbConnect(duckdb::duckdb())
duckdb_sales <- dplyr::copy_to(con, retail_sales, "retail_sales")
```

```{r}
#| label: nested-aggregate
#| must_error: true
#| eval: !expr has_duckdb

duckdb_sales |>
  summarize_with_margins(
    above_mean_share =
      sum(revenue[revenue > mean(revenue)]) / sum(revenue),
    .by = year,
    .grouping = rollup(region, store)
  ) |>
  collect()
```

`expand_with_margins()` is the way through. It emits one copy of each source
row for every grouping set, which turns the one summary pass into ordinary
rows you can take as many passes over as the calculation needs — a window
function for the group's mean, then an aggregate over the comparison. It stays
lazy throughout:

```{r}
#| eval: !expr has_duckdb

duckdb_sales |>
  expand_with_margins(
    .by = year,
    .grouping = rollup(region, store),
    .id = "set"
  ) |>
  group_by(year, region, store, set) |>
  mutate(set_mean = mean(revenue, na.rm = TRUE)) |>
  summarize(
    above_mean_share =
      sum(ifelse(revenue > set_mean, revenue, 0), na.rm = TRUE) /
      sum(revenue, na.rm = TRUE),
    .groups = "drop"
  ) |>
  collect()
```

```{r}
#| include: false
#| eval: !expr has_duckdb

DBI::dbDisconnect(con, shutdown = TRUE)
```

The values match the local calculation above. The cost is in the row count:
expansion produces one copy of the input per grouping set, so a plan with many
sets multiplies the rows the database has to scan. Reach for this when the
aggregation pass genuinely cannot express the calculation, not to avoid
writing a summary.

[Database and lazy
backends](https://sayuks.github.io/marginplyr/vignettes/database_backends.html)
shows the SQL `expand_with_margins()` generates on native and fallback
backends.
